---
title: "PurpleAir QC Comparison"
author: "Mazama Science"
date: "6/22/2020"
output: html_document
params: 
  label: 'SCAP_36'
  startdate: 20190701
  enddate: 20190708
---

Ashley's list of sensors to explore:

```
Below is a list of sensors and timeframes I was using to explore the results of 
applying different QC algorithms. I was primarily focused on examples where we 
see moderate to severe sensor noise from one or both channels (as opposed to more 
extreme/obvious sensor malfunctions, such as one channel reading zero).
 
Sensor: “SCNP_17”, startdate = 20200501, enddate = 20200520
Sensor: “CARB_SMOKE_SLO_CDF”, startdate = 20200501, enddate = 20200520
Sensor: “1105 Trail View Place, Nipomo, CA”, startdate = 20200501, enddate = 20200520
Sensor: "CARB_Smoke_SBCAPCD_Santa Maria", startdate = 20200501, enddate = 20200520
Sensor: "CARB_SMOKE_SLO_CDF", startdate = 20200501, enddate = 20200520
Sensor: "SCAP_19", startdate = 20200501, enddate = 20200520
Sensor: "SCTV_31", startdate = 20200501, enddate = 20200520
Sensor: "RUSD_1 (Arlington HS)", startdate = 20200516, enddate = 20200521
 
GitHub IO examples:
Sensor: “SCNP_20”, startdate = 20190411, enddate = 20190521
Sensor: "SCAP_46", startdate = 20190701, enddate = 20190710
 
GitHub IO example (Australian wildfire):
Sensor: “Downer”, startdate = 20191225, enddate = 20200110
```

-----

```{r setup, include = TRUE}
library(AirSensor)
AirSensor::initializeMazamaSpatialUtils()
setArchiveBaseUrl("http://data.mazamascience.com/PurpleAir/v1") # SCAQMD sensors 

knitr::opts_chunk$set(fig.width = 10, fig.height = 5)

```

```{r load, include = TRUE, warning = FALSE}
pas <- pas_load(archival = TRUE)
pat <- pat_createNew(
  label = params$label,
  pas = pas,
  startdate = params$startdate, 
  enddate = params$enddate
)
```

## Review PAT data

### Sensor Multi-plot

```{r multiplot, warning = FALSE}
pat_multiPlot(pat) 
```

### Sensor State of Health (Daily)

PurpleAir daily state-of-health (`pat_dailySoH()`) applies six algorithms:
- `PurpleAirSoH_dailyPctDC` 
- `PurpleAirSoH_dailyPctReporting`
- `PurpleAirSoH_dailyPctValid`
- `PurpleAirSoH_dailyMetFit`
- `PurpleAirSoH_dailyABFit`
- `PurpleAirSoH_dailyABtTest`

```{r soh, warning = FALSE, fig.height = 10}
pat_dailySoHPlot(pat)
```

-----

## Review QC Algorithms

See `?pat_dailySoH`, `?PurpleAirSoH_dailyPctDC`, `?PurpleAirSoH_dailyPctReporting`, `?PurpleAirSoH_dailyPctValid`, `?PurpleAirSoH_dailyPctValid`, `?PurpleAirSoH_dailyMetFit`, `?PurpleAirSoH_dailyABFit`, and `?PurpleAirSoH_dailyABtTest` for more details. 

### PurpleAir Quality-Control 00

Creates a `pm25` timeseries by averaging aggregated data from the A and B channels and applying the following QC logic:

- Create pm25 by averaging the A and B channel aggregation means
- Invalidate data where:  (min_count < 20)

```{r qc_00, warning = FALSE}
pat_monitorComparison(pat, FUN = AirSensor::PurpleAirQC_hourly_AB_00)
```

See `?AirSensor::PurpleAirQC_hourly_AB_00` for more details regarding the quality-control algorithm.

### PurpleAir Quality-Control 01

Creates a `pm25` timeseries by averaging aggregated data from the A and B channels and applying the following QC logic:

- Create pm25 by averaging the A and B channel aggregation means
- Invalidate data where:  (min_count < 20)
- Invalidate data where:  (p-value < 1e-4) & (mean_diff > 10)
- Invalidate data where:  (pm25 < 100) & (mean_diff > 20)

```{r qc_01, warning = FALSE}
pat_monitorComparison(pat, FUN = AirSensor::PurpleAirQC_hourly_AB_01)
```

See `?AirSensor::PurpleAirQC_hourly_AB_01` for more details regarding the quality-control algorithm.

### PurpleAir Quality-Control 02

Creates a `pm25` timeseries by averaging aggregated data from the A and B channels and applying the following QC logic:

- Create pm25 by averaging the A and B channel aggregation means
- Invalidate data where:  (min_count < 20)
- Invalidate data where:  (A/B hourly median-absolute-deviation (MAD) > 3)
- Invalidate data where:  (A/B hourly percent difference (pct_diff) > 0.5)

```{r qc_02, warning = FALSE}
pat_monitorComparison(pat, FUN = AirSensor::PurpleAirQC_hourly_AB_02)
```

See `?AirSensor::PurpleAirQC_hourly_AB_02` for more details regarding the quality-control algorithm.

### PurpleAir Quality-Control 03

Creates a `pm25` timeseries by averaging aggregated data from the A and B channels and applying the following QC logic:

- Create pm25 by averaging the A and B channel aggregation means
- Invalidate data where:  (min_count < 20)
- Invalidate data where:  (A/B hourly difference > 5 AND A/B hourly percent difference > 70%)
- Invalidate data where:  (A/B hourly data recovery < 90%)


```{r qc_03, warning = FALSE, fig.width = 10}
pat_monitorComparison(pat, FUN = AirSensor::PurpleAirQC_hourly_AB_03)
```

See `?AirSensor::PurpleAirQC_hourly_AB_03` for more details regarding the quality-control algorithm.

-----

## Compare QC Algorithms

```{r qc_comparison, fig.height = 10}
qc_00 <- pat_createAirSensor(pat, FUN = PurpleAirQC_hourly_AB_00)
qc_01 <- pat_createAirSensor(pat, FUN = PurpleAirQC_hourly_AB_01)
qc_02 <- pat_createAirSensor(pat, FUN = PurpleAirQC_hourly_AB_02)
qc_03 <- pat_createAirSensor(pat, FUN = PurpleAirQC_hourly_AB_03)

# Using baseplot for a custom graphic
layout(matrix(seq(3)))

PWFSLSmoke::monitor_timeseriesPlot(qc_00, pch = 16, col = 'red')
PWFSLSmoke::monitor_timeseriesPlot(qc_01, pch = 16, col = 'black', add = TRUE)
title("QC_hourly_AB_01")

PWFSLSmoke::monitor_timeseriesPlot(qc_00, pch = 16, col = 'red')
PWFSLSmoke::monitor_timeseriesPlot(qc_02, pch = 16, col = 'black', add = TRUE)
title("QC_hourly_AB_02")

PWFSLSmoke::monitor_timeseriesPlot(qc_00, pch = 16, col = 'red')
PWFSLSmoke::monitor_timeseriesPlot(qc_03, pch = 16, col = 'black', add = TRUE)
title("QC_hourly_AB_03")

layout(1)
```

-----

_Mazama Science_