---
title: "Downloading From ThingSpeak"
output: html_document
---

```{r load-libraries, echo=FALSE}
library(AirSensor)
library(ggplot2)
library(microbenchmark)
```

## Download Speed

Using the 'microbenchmark' R package we can determine the average processing 
time for different ThingSpeak data requests. Let's first incrementally test the
speed of 1 hour to 24 hour downloads. We will be using the 'Seattle' PurpleAir 
monitor for all requests and each test will be run 20 times.

```{r hour-benchmark-tests, echo=FALSE}
pas <- AirSensor::pas_load()

hour_speeds <- microbenchmark::microbenchmark(
  hrs_01 = downloadParseTimeseriesData(pas, label = "Seattle", 
                                       startdate = "2018-06-01 0", 
                                       enddate   = "2018-06-01 1"),
  hrs_02 = downloadParseTimeseriesData(pas, label = "Seattle", 
                                       startdate = "2018-06-01 0", 
                                       enddate   = "2018-06-01 2"),
  hrs_03 = downloadParseTimeseriesData(pas, label = "Seattle", 
                                       startdate = "2018-06-01 0", 
                                       enddate   = "2018-06-01 3"),
  hrs_04 = downloadParseTimeseriesData(pas, label = "Seattle", 
                                       startdate = "2018-06-01 0", 
                                       enddate   = "2018-06-01 4"),
  hrs_05 = downloadParseTimeseriesData(pas, label = "Seattle", 
                                       startdate = "2018-06-01 0", 
                                       enddate   = "2018-06-01 5"),
  hrs_06 = downloadParseTimeseriesData(pas, label = "Seattle", 
                                       startdate = "2018-06-01 0", 
                                       enddate   = "2018-06-01 6"),
  hrs_07 = downloadParseTimeseriesData(pas, label = "Seattle", 
                                       startdate = "2018-06-01 0", 
                                       enddate   = "2018-06-01 7"),
  hrs_08 = downloadParseTimeseriesData(pas, label = "Seattle", 
                                       startdate = "2018-06-01 0", 
                                       enddate   = "2018-06-01 8"),
  hrs_09 = downloadParseTimeseriesData(pas, label = "Seattle", 
                                       startdate = "2018-06-01 0", 
                                       enddate   = "2018-06-01 9"),
  hrs_10 = downloadParseTimeseriesData(pas, label = "Seattle", 
                                       startdate = "2018-06-01 0", 
                                       enddate   = "2018-06-01 10"),
  hrs_11 = downloadParseTimeseriesData(pas, label = "Seattle", 
                                       startdate = "2018-06-01 0", 
                                       enddate   = "2018-06-01 11"),
  hrs_12 = downloadParseTimeseriesData(pas, label = "Seattle", 
                                       startdate = "2018-06-01 0", 
                                       enddate   = "2018-06-01 12"),
  hrs_13 = downloadParseTimeseriesData(  pas, label = "Seattle", 
                                         startdate = "2018-06-01 0", 
                                         enddate   = "2018-06-01 13"),
  hrs_14 = downloadParseTimeseriesData( pas, label = "Seattle", 
                                         startdate = "2018-06-01 0", 
                                         enddate   = "2018-06-01 14"),
  hrs_15 = downloadParseTimeseriesData(pas, label = "Seattle", 
                                         startdate = "2018-06-01 0", 
                                         enddate   = "2018-06-01 15"),
  hrs_16 = downloadParseTimeseriesData(pas, label = "Seattle", 
                                         startdate = "2018-06-01 0", 
                                         enddate   = "2018-06-01 16"),
  hrs_17 = downloadParseTimeseriesData(pas, label = "Seattle", 
                                         startdate = "2018-06-01 0", 
                                         enddate   = "2018-06-01 17"),
  hrs_18 = downloadParseTimeseriesData( pas, label = "Seattle", 
                                         startdate = "2018-06-01 0", 
                                         enddate   = "2018-06-01 18"),
  hrs_19 = downloadParseTimeseriesData(pas, label = "Seattle", 
                                         startdate = "2018-06-01 0", 
                                         enddate   = "2018-06-01 19"),
  hrs_20 = downloadParseTimeseriesData(pas, label = "Seattle", 
                                         startdate = "2018-06-01 0", 
                                         enddate   = "2018-06-01 20"),
  hrs_21 = downloadParseTimeseriesData(pas, label = "Seattle", 
                                         startdate = "2018-06-01 0", 
                                         enddate   = "2018-06-01 21"),
  hrs_22 = downloadParseTimeseriesData( pas, label = "Seattle", 
                                         startdate = "2018-06-01 0", 
                                         enddate   = "2018-06-01 22"),
  hrs_23 = downloadParseTimeseriesData(  pas, label = "Seattle", 
                                         startdate = "2018-06-01 0", 
                                         enddate   = "2018-06-01 23"),
  hrs_24 = downloadParseTimeseriesData( pas, label = "Seattle", 
                                         startdate = "2018-06-01 0", 
                                         enddate   = "2018-06-02 0"),
  times = 20
)

hour_speeds
```

We can visually compare these benchmarks using the base R boxplot. ggplot2 also
provides a nice violin plot on calling autoplot.

```{r hour-benchmark-plots, message=FALSE, warning=FALSE, error=FALSE}
# Base R boxplot
boxplot(hour_speeds, unit = "s", log = FALSE, horizontal = TRUE,
        xlab = "Request", ylab = "Loadtime (Seconds)")

# ggplot autoplot
ggplot2::autoplot(hour_speeds, log = FALSE)
```

## Download Limit

In the past, ThingSpeak has limited downloads to about 8,200 records per 
request. Let's see if that's still the case by requesting a relatively lengthy
timeframe, say, two weeks, and then counting the total number of readings 
received.

```{r limit-test}
two_weeks <- downloadParseTimeseriesData(pas, label = "Seattle", 
                                         startdate = "2018-06-01 0", 
                                         enddate   = "2018-06-15 0")
records_count <- nrow(two_weeks$data)
first_record <- strftime(two_weeks$data[1,]$datetime, "%Y-%m-%d")
last_record <- strftime(two_weeks$data[records_count,]$datetime, "%Y-%m-%d")
```

Now we get `r records_count` entries in total. Although we specified to go from
`2018-06-01` to just before `2018-06-15`, the data we received ranges from 
`r first_record` to `r last_record`.