---
title: "An Introduction to PurpleAir Synoptic Data"
author: "Mazama Science"
date: "2019-04-19"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An Introduction to PurpleAir Synoptic Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(fig.width=7, fig.height=5)

library(AirSensor)
library(dplyr)
library(ggplot2)

# NOTE: Use example PAS data for vignettes to avoid long wait-times
data("example_pas")
pas <- example_pas
```

_Synoptic data_ provides a synopsis - a comprehensive and nearly instantaneous 
picture of the state of the atmosphere. This vignette demonstrates an example 
workflow for exploring air quality synoptic data using the AirSensor 
R package and data captured by [PurpleAir](https://www.purpleair.com/) air 
quality sensors.

### Loading PurpleAir Data
PurpleAir sensor readings are uploaded to the cloud every 80 seconds or so 
where they are stored for download and display on the PurpleAir website. In 
order to capture and download the latest synoptic dataset, begin by importing 
the required Mazama Science packages, setting the location of pre-generated
archive files and loading the most recent PurpleAir Synoptic ("`pas`") data.   

```{r, eval=FALSE}
library(AirSensor)
library(dplyr)
library(ggplot2)

setArchiveBaseUrl("http://smoke.mazamascience.com/data/PurpleAir")

pas <- pas_load()
```

The `pas_load()` function is a wrapper function that will: 

1. Download a raw dataset of the entire PurpleAir network that includes both 
metadata and recent PM2.5 averages for each deployed sensor across the globe. 
See `downloadParseSynopticData()` for more info.

2. Enhance the raw dataset by replacing variables with more consistent and 
human readable names, adds spatial metadata for each sensor, and lists the 
nearest official air quality monitor. For a more in depth explanation, see 
`enhanceSynopticData()`.

### PurpleAir Synoptic Data
The `pas` dataset contains 35 columns, and each row corresponds to different 
PurpleAir sensors. For the scope of this example we will focus on the columns
labeled `stateCode`, `pm25_*`, `humidity`, `pressure`, `temperature`, and 
`pwfsl_closestDistance`. 

The `pm25_*` column is short-hand for all the columns containing PM2.5 data. 
The extended names are fairly obvious; the `pm25_current` column contains data for 
the most recent air data, `pm25_10min` contains a 10-minute average of PM2.5 
data, and so on. 

```{r}
pas %>% 
  select(starts_with("pm25_")) %>% 
  head(5)
``` 

### Mapping `pas` PM2.5 data

To visually explore a region, we can use our `pas` data with the `pas_leaflet()`
function to plot an interactive [leaflet](https://leafletjs.com/) map.
By default, `pas_leaflet()` will map the coordinates of each PurpleAir sensor 
and the hourly PM2.5 data. Clicking on a sensor will show sensor metadata.

```{r}
pas %>% 
  pas_leaflet()
```

If we want to narrow our selection, for example to California, and look at which 
locations have a moderate to unhealthy 6-hour average air quality rating we can 
create a script like:

```{r CAleaflet}
pas %>% 
  filter(stateCode == 'CA') %>% 
  filter(pm25_6hr >= 25.0) %>% 
  pas_leaflet(parameter = "pm25_6hr")
```

This code pipes our `pas` data into `dplyr::filter()` where we can set our 
selection criteria. A `stateCode` is the ISO 3166-2 state name abbreviation, 
hence we limit our filter to only accept `stateCode == 'CA'`. The `pm25_6hr` 
selection, as mentioned above, selects the 6-hour average and restricts the 
sensors to those whose PM2.5 6-hour average is above 25.0. Be sure to include 
the `"pm25_6hr"` parameter in `pas_leaflet()`.  

### Mapping `pas` auxillary data

We can also explore and utilize other PurpleAir sensor data. Check the 
`pas_leaflet()` documentation for all supported parameters.

Here is an example of humidity data captured from PurpleAir sensors across the 
state of California.
```{r}
pas %>% 
  filter(stateCode == "CA") %>% 
  pas_leaflet(parameter = "humidity")
```

### Exploring PurpleAir data

Because the `pas` object is a dataframe, we can use functionality from the 
**dplyr** package to manipulate the data.

Below, we show how to find out which states have the highest _percentage_ of 
PurpleAir sensors that are reporting a weekly average of >20 ug/m3. 

```{r, warning=FALSE}
sensorCount <- 
  pas %>% 
  count(stateCode, name="sensorCount", sort = TRUE)

gt20Count <- 
  pas %>%
  select(stateCode, pm25_1week) %>% 
  filter(pm25_1week > 20.0) %>%
  count(stateCode, sort = TRUE, name = "gt20Count")

df <-
  left_join(gt20Count, sensorCount, by = "stateCode") %>%
  mutate(gt20Fraction = gt20Count / sensorCount) %>%
  arrange(desc(gt20Fraction)) %>%
  top_n(15)
```

From our `df` data frame, we can use `ggplot2` to visualize our results. 

```{r, warning=FALSE}
df %>% 
  ggplot(aes(x = reorder(stateCode, gt20Fraction), y = gt20Fraction)) + 
  coord_flip() + 
  geom_col(aes(fill = gt20Fraction)) + 
  scale_y_continuous(labels = scales::percent) + 
  labs(x = "State", y = "Percent of PA sensors above 20 \u00B5g/m3") + 
  theme(legend.position = "none")
```

As you can see, the AirSensor package combined with data provided by 
PurpleAir offers a powerful system for regional air-quality assessment. 

----

_Mazama Science_


