---
title: "Introduction to working with synoptic Purple Air data using MazamaPurpleAir"
author: "Ruby Fore"
date: "1/30/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=7, fig.height=5)
```


## 'Spatially enhanced metadata' - looking inside `pas_load`

The function `pas_load` does its job by calling two other functions in the **MazamaPurpleAir** package: `enhanceSynopticData` and `downloadParseSynopticData`. To better understand the `pa_synoptic` object returned by `pas_load`, a closer look at these two functions is merited. 

First, a quick definition of synoptic. Synoptic means 'relating to or displaying conditions (as of the atmosphere or weather) as they exist simultaneously over a broad area'. It is aptly used here as the counterpoint to timeseries data - synoptic data is from all sensors at one timepoint, timeseries data is from one sensor over multiple timepoints. 

# The downloadParseSynopticData() function

As suggested by its name, `downloadParseSynopticData()` acquires and digests data from the entire Purple Air network, (**question: over what time frame?**) and formats it into an R `data.frame` from the original JSON. Let's examine a  dataset that is the raw data downloaded using `downloadParseSynopticData` alone, not in the `pas_load` function.   

The first step is loading the **MazamaPurpleAir** package. You can then either use a built in dataset, `pas_raw`, or download your data from the site yourself. The `pas_raw` data set was created by an identical command to the one below under "code to download new data".  

```{r loading-packages-and-data, message=FALSE, warnings='suppress'}
# loading required package
library(MazamaPurpleAir)

# using pre-loaded data
data("pas_raw")

# code to download new data 
# pas_raw <- downloadParseSynopticData()
```

Let's look at the column/variable names in this dataset. 
```{r names}
names(pas_raw)
```
 
 
The first thirteen look pretty good, but 23-29 (v, v1, *etc*) are not human-readable. It would be nice to have a descriptive name for these columns. This is one enhancement accomplished by `enhanceSynopticData`.

# The enhanceSynopticData() function





## using dplyr to filter based on different criteria
## creating interactive maps with pas_leaflet()