---
title: "BikeSGV Story"
output:
  html_document:
    df_print: paged
---

Using just few of AirSensor's synoptic and time series functions, we can 
investigate something fishy going on with one particular air sensor in 
northwest Pasadena.

Let's pull it up by its sensor label, *BikeSGV - West Pasadena*, from a loaded 
list of US PurpleAir sensors:

```{r setup}
library(AirSensor)

setArchiveBaseUrl("http://smoke.mazamascience.com/data/PurpleAir")

pas <- pas_load()
pas_bikesgv <- pas_filter(pas, label == "BikeSGV - West Pasadena")
pas_leaflet(pas_bikesgv)
```

While this surface level view may seem normal at first, the sensor's past
measurements happen to display something very strange. We can gather these 
readings by loading a PurpleAir Timeseries *pat* object using the same label:

```{r load_pat_and_multiplot, warning=FALSE}
pat_bikesgv <- pat_createNew(pas, "BikeSGV - West Pasadena", 
                            startdate="2019-06-18", enddate="2019-06-25")
pat_multiplot(pat_bikesgv)
```

This sensor's A and B PM sensors appear to be telling us two very different 
stories! Let's ignore temperature and humidity for now and zoom in on just the 
PM readings by setting *plottype*:

```{r plot_only_pm25, warning=FALSE}
pat_multiplot(pat_bikesgv, plottype = "pm25_over")
```

While the A channel PM 2.5 measurements oscillate over time in a fairly natural 
manner, they also all measure impossibly high: Around 4000 μg/m³ when 
anything over 100 μg/m³ on the 
[AQI](https://www.airnow.gov/index.cfm?action=aqibasics.aqi) can be dangerous!
This also skews the plot enough to make channel B look practically flat, so 
let's separate them and take a look at B on its own scale to see if it's 
readings are any more reasonable:

```{r plot_only_pm25b, warning=FALSE}
pm25_b <- pat_bikesgv$data[, c("datetime", "pm25_B")]
plot(pm25_b, type = "p", cex = 0.6, pch=15, col=adjustcolor("black", 0.2),
     main = "PM 2.5 Channel B", xlab = "2019", ylab = "PM 2.5 (μg/m³)")
```

B channel does look a bit more normal. There are a few spikes but they never 
reach the ridiculous levels of channel A. So then what *is* the issue with the A 
channel? Is there any relationship between the channels at all? Let's generate a
linear fit between them and see if they might have a common scale factor:

```{r channel_fit_plot, warning=FALSE, message=FALSE, error=FALSE}
pat_internalFit(pat_bikesgv)
```

It looks like the A and B channels aren't linearly correlated at all. This is 
odd since the two sensors are measuring the same air and should at least give
somewhat similar readings. Maybe comparing more sensors could give us some 
insight. We can check them all against each other by using a scatterplot:

```{r all_sensors_scatterplot, warning=FALSE, message=FALSE}
pat_scatterplot(pat_bikesgv)
```

Now we're getting somewhere! While the correlation plot between the two channels
is a shapeless cloud, channel A seems to have a very well defined relationship 
with temperature and humidity while channel B has no similarities with them at 
all.

Just for reference, this is the scatterplot for a different sensor, only a few
hundred feet away, that is not experiencing this issue:

```{r normal_scatterplot, warning=FALSE, message=FALSE}
pat_normal <- pat_createNew(pas, "AQMD_NASA_33", 
                            startdate="2019-06-18", enddate="2019-06-25")
pat_scatterplot(pat_normal)
```

Notice that at this site, both PM 2.5 sensors share a proper linear relationship
and neither channel has much correlation with the auxillary sensors.

The BikeSGV sensor is quite the opposite. Let's take a closer look at how its 
A channel is related to temperature by generating a linear model: 

```{r pm25a_and_temp, warning=FALSE}
pat_data <- pat_bikesgv$data
model_temp <- lm(temperature ~ pm25_A, data=pat_data)

par(mfrow = c(2, 2), las = 1)
plot(model_temp)
summary(model_temp)
```

That's a pretty clear correlation! Now how closely would the channel A plot 
match that of temperature if we apply the fit model coefficients?

```{r plot_fitted_pm25a_to_temp}
x <- pat_data$datetime
y <- pat_data$temperature
temp_fitted_pm25a <- predict.lm(model_temp, newdata=pat_data)

par(las = 1)
plot(x, temp_fitted_pm25a, 
     pch = 15, cex = 0.9, col=adjustcolor("black", 0.2),
     main = "Fitting PM2.5_A to Temperature",
     xlab = "2019", ylab = "Temperature (F)")
lines(x, y, lwd = 4, col=adjustcolor("tomato", 0.7))
legend(x = "bottomright", legend = c("Temp", "Fitted ChA"), 
       col = c("tomato", "black"), bg = adjustcolor("lightgray", 0.3), 
       lwd = 4, box.lty = 0, inset = 0.025)
```

Wow! Although it's a bit noisy, the A channel data definitely matches the 
general shape of the temperature plot! Let's see if the same holds when 
comparing to humidity:

```{r plot_fitted_pm25a_to_humidity}
model_humid <- lm(humidity ~ pm25_A, data=pat_data)
x <- pat_data$datetime
y <- pat_data$humidity
hum_fitted_pm25a <- predict.lm(model_humid, newdata=pat_data)

par(las = 1)
plot(x, hum_fitted_pm25a, 
     pch = 15, cex = 0.9, col=adjustcolor("black", 0.2),
     main = "Scaling PM2.5_A to Humidity",
     xlab = "2019", ylab = "Humidity (RH)")
lines(x, y, lwd = 4, col=adjustcolor("cornflowerblue", 0.7))
legend(x = "bottomright", legend = c("Humidity", "Fitted ChA"), 
       col = c("cornflowerblue", "black"), bg = adjustcolor("lightgray", 0.3), 
       lwd = 4, box.lty = 0, inset = 0.025)
```

Though maybe not as definite as the temperature comparison, the channel A 
readings also match the trend of the humidity data. Just to be complete, let's
try the same comparison process for channel B and see if we find a relationship
there too:

```{r plot_fitted_pm25b_to_temp}
model_temp_b <- lm(temperature ~ pm25_B, data=pat_data)
x <- pat_data$datetime
y <- pat_data$temperature
temp_fitted_pm25b <- predict.lm(model_temp_b, newdata=pat_data)

par(las = 1)
plot(x, y, 
     col=adjustcolor("black", 0.0),
     main = "Fitting PM2.5_B to Temperature",
     xlab = "2019", ylab = "Temperature (F)")
points(x, temp_fitted_pm25b, 
       pch = 15, cex = 0.9, col=adjustcolor("black", 0.2))
lines(x, y, lwd = 4, col=adjustcolor("tomato", 0.8))
legend(x = "bottomright", legend = c("Temp", "Fitted ChB"), 
       col = c("tomato", "black"), bg = adjustcolor("lightgray", 0.3), 
       lwd = 4, box.lty = 0, inset = 0.025)
```

Hmm, not quite for this sensor.
So what is happening with channel A? It almost seems like the
temperature or humidity readings have been leaking into the PM 2.5 channel A 
sensor and overriding its measurements.

So far though we've only looked at a week's worth of questionable data. Has this
sensor always been acting this way? Lets load and view readings from earlier 
this year:

```{r april_dygraph, warning=FALSE}
pat_april <- pat_createNew(pas, "BikeSGV - West Pasadena", 
                           startdate="2019-04-05", enddate="2019-04-22")
pat_dygraph(pat_april)
```

Well that answers that question! Before the evening of April 18th, 2019, both 
channels seemed to agree with each other for the most part. Around 7:00 PM that 
day though, channel A suddenly switched to the strange state it's currently 
stuck in. Channel A also appeared to have some issues between April 7th and 9th 
which yielded readings in the same 3000-4000 μg/m³ range, but returned to normal
for about a week.

Channel B seemed unaffected by anything the whole time though. Let's see if 
the auxilary sensors experienced anything unusual:

```{r transition_multiplot, warning=FALSE}
pat_multiplot(pat_april)
```

It seems that only the channel A sensor was affected at this time. The reason
why is still unknown, but at the very least we now have a good idea of where 
this sensor is getting its readings and when this change occurred.

THE END
