---
title: "Tutorial: Creating A Local Archive For PurpleAir Sensors"
author: "Mazama Science"
date: "2021-03-01"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 5)
```

This article illustrates how to create a local archive of pre-generated data 
files for PurpleAir sensors, in order to use the **[AirSensors](https://mazamascience.github.io/AirSensor/index.html)** package off-line. 
For the purpuse of this tutorial, we are going to create a local archive
for the Methow Valley's sensors by loading pre-generated 
PurpleAir synoptic and time searies files from an [internal archive base url](http://data.mazamascience.com/PurpleAir/v1/) mantained by the AirSensor 
package. This is the fastest way for obtaning sensors data but, alternatively, 
you could download the data directly from the PurpleAir website. 
More info on this alternative method is available in the [PurpleAir Synoptic Data](https://mazamascience.github.io/AirSensor/articles/articles/pas_introduction.html) and 
[PurpleAir Time Series Data](https://mazamascience.github.io/AirSensor/articles/articles/pat_introduction.html) articles. 


## Section 1: loading pre-generated files to create a *flat* archive 

**Step 1** Load the `AirSensor` and `dplyr` libraries. You may need to install 
other necessary packages first (see [AirSensor R Packeage](https://mazamascience.github.io/AirSensor/index.html) for more details).

```{r, libraries, warning = FALSE, message=FALSE}
library(AirSensor)
library(dplyr)
```

**Step 2** Set the archive base directory to NULL. This step is necessary only 
if you previously set a base directory, since it would take precedence over the 
base url we are going to set in Step 3.

```{r, setArchiveBaseDir, warning = FALSE, message=FALSE}
setArchiveBaseDir(NULL)
```

**Step 3** Set the archive base url to http://data.mazamascience.com/PurpleAir/v1/. 
```{r, setArchiveBaseUrl, warning = FALSE, message=FALSE}
setArchiveBaseUrl("http://data.mazamascience.com/PurpleAir/v1/")
```

**Step 4** Create a `pas` (PurpleAir Synoptic) object for
the Methow Valley'sensors. The `pas` object contains all the sensor's data 
and metadata, including latitude and longitude which allow us to easily visualize
the sensors on a map. You can then click the sensors on the map to see the most 
important metadata and data.
```{r, pas object and map, warning = FALSE, message=FALSE, results='hide'}
mvcaa <-
  pas_load() %>% # load the most recent archived 'pas'(PurpleAir Synoptic)
  pas_filter(stringr::str_detect(label, "MV Clean Air Ambassador"))
```
```{r, dynamic map}
pas_leaflet(mvcaa) # dynamic map of the sensors.
```
<br>
**Step 5** Create a list of unique time series identifiers or `DeviceDeploymentIDs`.
```{r}
ID <- pas_getDeviceDeploymentIDs(pas = mvcaa)
```

If you only want to obtain a list of unique time series identifiers, skip Step 4
and 5, and follow the example below.

```{r, unique time series identifiers, eval=FALSE}
ID <- pas_load() %>%
  pas_getDeviceDeploymentIDs("MV Clean Air Ambassador")
```
<br>
Now that we have our list of unique time series identifiers, we can load the PurpuleAir time series data (heareafter `pat`) to build our local archive. 
Here we have two options: 1) build a *flat* archive 
where we simply store files; 2) build an archive using the *well-known* structure
(aka archive base url structure). The latter will allow you to use 
AirSensors `~load` functions for further off-line analysis, including:

* `pas_load()`

* `pat_load()`

* `pat_loadLatest()`

* `pat_loadMonth()`

* `sensor_load()`

* `sensor_loadLatest()`

* `sensor_loadMonth()`

If you choose option 1), got to Step 6; if you choose option 2) jump to Section 
2.

**Step 6** Load and save `pas` and `pat` files in a *flat* archive. 
```{r, load pat data for September 2020, eval=FALSE}

my_archive <- "~/my_archive" # type here your the directory where you want to 
                             # save the files 

# Load and save pat files for September 2020
for (id in ID){
  filename = paste0(my_archive, "/", id, ".rda")
  pat = pat_createNew(id = id,
                      pas = mvcaa,
                      startdate = 20200901, 
                      enddate = 20200930)
  save(pat, file = filename)
} 

# save the pas object created in Step 4 in your flat archive as an .rda file
save(mvcaa, file = paste0(my_archive, "/mvcaa.rda"))

# to load files from your flat archive you can use the get(load()) function 
mvcaa <- get(load(paste0(my_archive,"/mvcaa.rda")))

```


## Section 2: loading pre-generated files and saving them using the *well-known* archive structure. 

At this point you have completed Step 1 through 5 of Section 1, thus you have
the `pas` data frame (`mvcaa`) and the `ID` list ready in your R environemnt 
and you can proceed with the next steps.

**Step 1** Create a local archive using the *well-known* archive structure:

* For `pas` files, the directory must have the following structure: ~pas/year.

* For `pat` files, the directory must have the following structure: ~pat/year/month.

* It is also critical to name the files as follow: pas_YYYYMMDD.rda or pat_id_YYYYMM.rda. 

* Examples: ~pas/2021/pas_20210301.rda or 
~pat/2020/09/pat_b56c0ef677852913_81495__202009.rda for `pas` or `pat` files respectively. 

```{r, build well-known archive structure, eval=FALSE}

# create the pas directory
pas_2021 <- "pas/2021"
dir.create(file.path(my_archive, pas_2021), recursive = TRUE)

# create the pat directory for September 2020
pat_202009 <- "pat/2020/09"
dir.create(file.path(my_archive, pat_202009), recursive = TRUE)

# save the new pat directory as an object (we'll need this in Step 3)
pat_202009 = paste0(my_archive,"/", pat_202009)
```

**Step 2** Save the `pas` object previously created (Step 4, Section 1) as an 
.rda file in your *well known* archive structure. 

```{r, save pas file - well known archive, eval=FALSE}
save(mvcaa, file = paste0(my_archive, "/pas/2021/pas_20210301.rda"))
```
*NOTE*: the date in the file name correspond to the day I loaded 
the data. I did it on March 1, 2021.


**Step 3** Load and save the `pat` objects as .rda files in your *well known* 
archive structure. You'll need the `ID` list previously created (Step 5, 
Section 1). 

*NOTE*: If one or more sensors do not have data available for the 
time range specified in the function, you'll get `Error in pat_filterDatetime(., startdate = dateRange[1], enddate = dateRange[2],  :Parameter 'pat' has no data`. 
This error message does not communicate which sensor has no data available, and 
we are working on improving it to make it more specific. 

```{r, load pat data for September 2020 - well known archive, eval=FALSE}

for (id in ID){
  filename = paste0(pat_202009, "/", "pat_", id, "_202009", ".rda")
  pat = pat_createNew(id = id,
                      pas = mvcaa,
                      startdate = 20200901,
                      enddate = 20200930)
  save(pat, file = filename)
}
```

**Step 4** Set your archive base directory and test if you can load the data 
from your archive using the AirSensor functions. 

*NOTE*: to load `pat` data it 
is important to *always* include the `startdate` and `enddate` arguments. 
For consitency, I suggest you to include the `pas` object as well.


```{r, load data with AirSensor functions, results='hide'}

# Set your archive base directory
setArchiveBaseDir(getwd())

# Check that your archive base directory is correct
getArchiveBaseDir()

# load the pas file (refer to Step 2)
pas <- pas_load(20210301) 

# load one of the pat files to make sure you can pull data from your well-known archive 
pat <- pat_load(id = "ab5dca99422f2c0d_13669",
         pas = pas,
         startdate = 20200901,
         enddate = 20200930, 
         timezone = "America/Los_Angeles")
```

**Step 5** Now that you have your well-known archive all set, have fun working off-line by using the [AirSensor functions](https://mazamascience.github.io/AirSensor/reference/index.html). 
In the example below, I'm plotting `pat` data using the `pat_multiplot()` and `pat_scatterPlotMatrix()`.

```{r, pat multiplot, warning=FALSE, message=FALSE}
pat_multiplot(pat)
```

```{r, pat scatterPlotMatrix, warning=FALSE, message=FALSE}
pat_scatterPlotMatrix(pat)
```


