---
title: "MVCAA Tutorial 2: Exploring PAT Data"
author: "Mazama Science"
date: "2021-03-02"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MVCAA Tutorial 2: Exploring PAT Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 5)
```

## Introduction

This tutorial demonstrates basic exploration of previously created 'pat' 
timeseries data. In order to run the code in this tutoral you must have 
followed the instructions in Tutorial 1 and created a directory with 'pas' 
and 'pat' data files for the Methow Valley. Target
audiences include grad students, researchers and any member of the public
concerned about air quality and comfortable working with R and RStudio.

Tutorials in this series include:

* [MVCAA Tutorial 1: Creating PAT Data](https://mazamascience.github.io/AirSensor/articles/articles/MVCAA_Tutorial_1.html)
* [MVCAA Tutorial 2: Examining PAT Data](https://mazamascience.github.io/AirSensor/articles/articles/MVCAA_Tutorial_2.html)
* [MVCAA Tutorial 3: Creating Airsensor Data](https://mazamascience.github.io/AirSensor/articles/articles/MVCAA_Tutorial_3.html)
* [MVCAA Tutorial 4: Exploring Airsensor Data](https://mazamascience.github.io/AirSensor/articles/articles/MVCAA_Tutorial_4.html)
* [MVCAA Tutorial 5: Building a Local Archive](https://mazamascience.github.io/AirSensor/articles/articles/MVCAA_Tutorial_5.html)
* [MVCAA Tutorial 6: Using a Local Archive](https://mazamascience.github.io/AirSensor/articles/articles/MVCAA_Tutorial_6.html)
* [MVCAA Tutorial 7: Methow Vallely Smoke](https://mazamascience.github.io/AirSensor/articles/articles/MVCAA_Tutorial_7.html)

## Goal

Our goal in this tutorial is to explore a variety of `pat_~()` functions available
in the **AirSensor** package and to explain their most important arguments and 
the interpretation of output graphics.

## [pat_dygraph](https://mazamascience.github.io/AirSensor/reference/pat_dygraph.html)
This function creates interactive graphs that will be displayed in RStudio's 'Viewer' tab.
```{r}
patList <- get(load(paste0(archiveDir,"/patList.rda")))
Balky_Hill <- patList[["ab5dca99422f2c0d_13669"]]
Balky_Hill_label <- Balky_Hill$meta$label #"MV Clean Air Ambassador @ Balky Hill"

# note: consider creating a loop to create pat objects to fill in the pat argument. 

pat_dygraph(
  pat = Balky_Hill, #from tutorial 1
  parameter = "pm25",
  sampleSize = 1000,
  title = NULL,
  xlab = "September 2020",
  ylab = NULL,
  tlim = NULL,
  rollPeriod = 6,
  showLegend = TRUE,
  colors = NULL,
  timezone = NULL
)

```


TODO

## pat_multiPlot()

TODO

## pat_scatterPlotMatrix()

TODO

## pat_outliers()

TODO

## pat_internalFit()

TODO

## pat_dailySoHPlot()

TODO

## pat_monitorComparison()

TODO

## pat_externalFit()

TODO



----

_Best of luck assessing air quality in your community!_
