---
title: "MVCAA Tutorial 2: Exploring PAT Data"
author: "Mazama Science"
date: "2021-03-16"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MVCAA Tutorial 2: Exploring PAT Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 5)
```

## Introduction

This tutorial demonstrates basic exploration of previously created 'pat' 
timeseries data. In order to run the code in this tutorial you must have 
followed the instructions in Tutorial 1 and created a directory with 'pas' 
and 'pat' data files for the Methow Valley. Target
audiences include grad students, researchers and any member of the public
concerned about air quality and comfortable working with R and RStudio.

Tutorials in this series include:

* [MVCAA Tutorial 1: Creating PAT Data](https://mazamascience.github.io/AirSensor/articles/articles/MVCAA_Tutorial_1.html)
* [MVCAA Tutorial 2: Exploring PAT Data](https://mazamascience.github.io/AirSensor/articles/articles/MVCAA_Tutorial_2.html)
* [MVCAA Tutorial 3: Creating Airsensor Data](https://mazamascience.github.io/AirSensor/articles/articles/MVCAA_Tutorial_3.html)
* [MVCAA Tutorial 4: Exploring Airsensor Data](https://mazamascience.github.io/AirSensor/articles/articles/MVCAA_Tutorial_4.html)
* [MVCAA Tutorial 5: Building a Local Archive](https://mazamascience.github.io/AirSensor/articles/articles/MVCAA_Tutorial_5.html)
* [MVCAA Tutorial 6: Using a Local Archive](https://mazamascience.github.io/AirSensor/articles/articles/MVCAA_Tutorial_6.html)
* [MVCAA Tutorial 7: Methow Vallely Smoke](https://mazamascience.github.io/AirSensor/articles/articles/MVCAA_Tutorial_7.html)

## Goal

Our goal in this tutorial is to explore a variety of `pat_~()` functions available
in the **AirSensor** package and to explain their most important arguments and 
the interpretation of output graphics.

## PurpleAir A & B Channels

PurpleAir (PA) sensors are unique in that they have two separate optical particle 
detectors — one labeled A and one labeled B.
All of the plotting functions, except `pat_monitorComparison()` and 
`pat_externalFit()`, only use data from a *single* PA sensor — with A and B data channels. 
When measurements from the A and B channels are largely in agreement,
the sensor is working properly. When they substantially differ, there is 
a problem. Moreover, when a sensor is properly functioning, correlation between 
the A and B channel (pm25_A:pm25_B) will be strongly positive, while correlation
between temperature and humidity will typically be strongly negative.

## Loading `pat` objects created in MVCAA Tutorial 1

```{r load-data, eval = TRUE, warning = FALSE, message = FALSE}
library(AirSensor)

# Use the default archiveDir unless it is already defined
if ( !exists("archiveDir") ) {
  archiveDir <- file.path("~/Data/MVCAA")
}

# Load pat data
patList <- get(load(file.path(archiveDir,"/patList.rda")))

# Print site ids and associated names
sapply(patList, function(x) { return(x$meta$label) }) 

# Pull out each sensor as a separate 'pat' object
Balky_Hill <- patList[["ab5dca99422f2c0d_13669"]]
Bush_School <- patList[["49215ad49d1a87e3_10188"]]
Liberty_School <- patList[["db5d6b3b79f5830e_39237"]] # good
McFarland_Creek <- patList[["f592adb5067ad9d3_13675"]]
Pine_Forest <- patList[["0cbfeb2ce4c1553c_13661"]]
Willowbrook_Farm <- patList[["f96deab8c29aa42b_10134"]]
Benson_Creek <- patList[["f6c44edd41c941c7_10182"]]
Gunn_Ranch <- patList[["f736fd3fb21fc4da_13667"]]
Lower_Studhorse <- patList[["4f19d256e1787973_10166"]]
Methow_Estates <- patList[["4a47b9252e16e558_15077"]] 
Beaver_Creek <- patList[["2e3b5ceea86a885b_10168"]]
Little_Cougar <- patList[["96b108298883ca47_64441"]]
```

## pat_dygraph

The [pat_dygraph()](https://mazamascience.github.io/AirSensor/reference/pat_dygraph.html)
function creates interactive graphs that will be displayed in RStudio's 'Viewer' tab.

Lets explore a couple of sensors. We will accept default settings for most of
the `pat_dygraph()` parameters but will apply use `sampleSize = 1000` and 
`rollPeriod = 6`. Reducing the `sampleSize` prevents the interactive viewer from
bogging down and increasing the `rollPeriod` results in smooth lines rather than
disconnected points.

Specifying the correct Olson `timezone` ensures that times are displayed in 
local time.

_NOTE:_ You can experiment with the `rollPeriod` by adjusting the number in the
dygraph viewer.

_NOTE:_ You can use the sliders underneath each plot to change the time period
or "brush" over the plot itself to restrict both X and Y ranges. Double-click
to return to full range.

### Liberty School

```{r Liberty-School, eval = TRUE, warning = FALSE, message = FALSE}
pat_dygraph(
  pat = Liberty_School, 
  sampleSize = 1000,
  xlab = "September 2020",
  rollPeriod = 6,
  timezone = "America/Los_Angeles"
)
```


### Willowbrook Farm

```{r Willowbrook-Farm, eval = TRUE, warning = FALSE, message = FALSE}
pat_dygraph(
  pat = Willowbrook_Farm,       
  sampleSize = 1000,
  xlab = "September 2020",
  rollPeriod = 6,
  timezone = "America/Los_Angeles"
)
```

For Liberty School, the A and B channel measurements are largely in agreement 
while for Willowbrook Farm, the channels are largely in disagreement. This could 
indicate a malfunctioning of the sensor, and it could be interesting to investigate
if this is an isolated event by exploring a larger time range. 

## pat_multiPlot 

The [pat_multiplot()](https://mazamascience.github.io/AirSensor/reference/pat_multiplot.html)
function allow us to quickly display our raw `pat` data using one of several
common plot types. Available `plottype` options include:

* "all" – pm25_A, pm25_B, temperature, humidity

* "pm25_a" – PM2.5 from channel A only

* "pm25_b" – PM2.5 from channel B only

* "pm25" – PM2.5 from channels A and B in separate plots

* "pm25_over" – PM2.5 from channels A and B in the same plot

* "aux" – auxiliary data (temperature, humidity)

For the purpose of this tutorial we'll use "all", "pm25_over", and "aux". 
As before, sub-sampling to only 1000 points speeds up plotting.

_NOTE:_ This plot carefully retains outliers in the A and B channels while 
sampling so that the overall appearance of the A and and B plots should remain 
the same regardless of the `sampleSize`.

```{r Liberty-School-all, eval = TRUE, warning = FALSE, message = FALSE}
# Liberty School multiplot - plottype = "all"
pat_multiPlot(
  pat = Liberty_School,
  plottype = "all",
  sampleSize = 1000
)

```

The "all" plot type offers a nice overview of all four variables. We can 
already start seeing an agreement between channel A and B data and an inverse relationship between humidity and temperature measurements. 

Let's give a closer look at the channels using the "pm25_over" plot type. 
```{r Liberty-School-pm25over, eval = TRUE, warning = FALSE, message = FALSE}
# Liberty School multiplot - plottype = "pm25_over"
pat_multiPlot(
  pat = Liberty_School,
  plottype = "pm25_over",
  sampleSize = 1000
)
```

Channels A and B show strong agreement, and we can improve our faith in this
assessment by increasing the sample size.  Let's try `sampleSize = 5000`. 

```{r Liberty-School-pm25over-n5000, eval = TRUE, warning = FALSE, message = FALSE}
# Balky Hill multiplot - plottype = "pm25_over", sampleSize = 5000
pat_multiPlot(
  pat = Liberty_School,
  plottype = "pm25_over",
  sampleSize = 5000
)
```

Again, channels A and B are largely in agreement. Purely 
based on visual inspection, we can assume that the Liberty School sensor is 
working correctly. However, more in-depth statistical analysis is discussed later 
in this tutorial. 

Let's give a closer look at the relationship between temperature and humidity 
by using the "aux" plot type.

```{r Liberty-School-aux, eval = TRUE, warning = FALSE, message = FALSE}
# Liberty School multiplot - plottype = "aux"
pat_multiPlot(
  pat = Balky_Hill,
  plottype = "aux",
  sampleSize = 5000
)
```

From the graph, we can see what looks like a negative correlation between the two 
variables, which supports the assumption that the sensor is functioning correctly. 

## pat_scatterPlotMatrix

The [pat_scatterPlotMatrix()](https://mazamascience.github.io/AirSensor/reference/pat_scatterPlotMatrix.html)
function returns a matrix of scatterplots and correlation values for the most
important variables in the `pat` object. Various sensor failure modes are often
evident as unexpected or missing correlations in this plot.

The list of available parameters include:

* datetime -- measurement time

* pm25_A -- A channel PM2.5 (ug/m3)

* pm25_B -- B channel PM2.5 (ug/m3)

* temperature -- temperature (F)

* humidity -- humidity (%)

Let's compare all of them for the Liberty School sensor.  

```{r pat-scatterPlotMatrix-Liberty-School, eval = TRUE, warning = FALSE, message = FALSE}
# Liberty School scatterPlotMatrix
pat_scatterPlotMatrix(
  pat = Liberty_School
)
```

As we can see from the matrix, using a sample size of 5000 observations 
for each variable, there is an excellent correlation of near 1 between 
channels A and B (pm25_A:pm25_B) and a strong negative correlation of around -0.84
between temperature and humidity. (We say "around" because we are using a sub-sample
of the entire population.) These results support the assumption that the 
sensor worked well throughout September 2020. 

## pat_outliers

The [pat_outliers()](https://mazamascience.github.io/AirSensor/reference/pat_outliers.html)
function detects outliers using a Median Average Deviation or "Hampel" filter.
This filter looks at a window of consecutive values and identifies those 
that are very far out in the tails of the distribution of values within the window. 
The filter window is then "rolled" over the entire length of the 
time series, much like a "rolling mean". 

`windowSize = 15` corresponds to a temporal window of approximately 30 minutes
given the current PurpleAir sampling frequency of 1/120 sec. This long enough
that we have a reasonable number of values from which to calculate sub-population statistics and
short enough that we would not expect radical changes in atmospheric conditions

The default threshold setting `thresholdMin = 8` identifies points that are
extremely unlikely to be part of a normal distribution and therefore very likely
to be outliers.

Specifying `replace = TRUE` will return a new pat object with outliers replaced
by the window median values. Using this technique, you can create a highly 
smoothed, artificial dataset by setting thresholdMin = 1 or lower (but always 
above zero).

You can create a new pat object and save it in your archive directory 
(already defined in Tutorial 1) as in the example below. 

```{r pat-outliers-Liberty-School, eval = TRUE, warning = FALSE, message = FALSE}
# create an outliers plot for Liberty School 
Liberty_School_clean <- pat_outliers(
  pat = Liberty_School,
  windowSize = 15,
  thresholdMin = 8,
  replace = TRUE,
  showPlot = TRUE
)

# Liberty_School_clean contains the replaced values and can be saved in your archive 
# directory.
save(Liberty_School_clean, file = file.path(archiveDir, "Liberty_School_clean.rda"))
```

The `pat_outlier()` function will always find some outliers, often insignificant 
ones. The plot above shows a few non-concerning outliers. Outliers of concern may be
associated with electronic "glitches" and appear far away from the main 
distribution as depicted in the example 
below (notice data around August 3, 2018). 

```{r AirSensor-example-patoutliers(), eval = TRUE, warning = FALSE, message = FALSE}
example_pat %>%
  pat_filterDate(20180801, 20180815) %>%
  pat_outliers()
```

## pat_internalFit

The [pat_internalFit()](https://mazamascience.github.io/AirSensor/reference/pat_internalFit.html)
function uses a linear model to fit data from channel B to data from 
channel A and returns an object of class "lm". If you assign the result to an object, you can 
then interrogate the linear model using the base R function `summary()`. 

The default setting of `showPlot = TRUE` will also generate a model fit plot.  

Let's have a look at a well functioning sensor and a not-so-well functioning
one. 

```{r lm-Liberty-School, eval = TRUE, warning = FALSE, message = FALSE}
# Liberty School linear model 
lm_Liberty <- pat_internalFit(
  pat = Liberty_School
)

summary(lm_Liberty)
```

The model fit plot shows that the model is slightly biased towards channel B 
(slope = 0.95) and that the linear model explains 99.9% of the data variability.
This supports the assumption that the Liberty School sensor functioned correctly
throughout most of September 2020.

Now lets take a look at our problem sensor:

```{r lm-Willowbrook-Farm, eval = TRUE, warning = FALSE, message = FALSE}
# Willowbrook Farm linear model
lm_Willowbrook <- pat_internalFit(
  pat = Willowbrook_Farm
)

summary(lm_Willowbrook)
```

The model fit plot shows that the model is biased towards channel A 
(slope = 1.5), as is also evident in the bottom graphic, indicating the 
malfunctioning of one or both Willowbrook Farm sensor's channels. 
The model explains only 92.2% of the data variability, reinforcing the assumption
that the sensor did not function properly in September, 2020. 

_But we knew something was wrong from a first glance at the plot!_

While statistical
calculations are important, we encourage everyone to use the **AirSensor** package 
to truly "look at" the data from these sensors.

## pat_dailySoHPlot

The [pat_dailySoHPlot()](https://mazamascience.github.io/AirSensor/reference/pat_dailySoHIndexPlot.html)
function is in a very different category from the other functions mentioned so
far. It works with daily "State-of-Health" metrics that attempt to identify days
when a sensor is performing well or poorly. The idea is that State-of-Health
metrics can be used as a sort of daily checkup so that people can identify 
sensors that are beginning to malfunction due to acute problems like disconnected
wires or due to gradual deterioration associated with old age.


The `pat_dailySoHPlot()` function plots a subset of the most useful State of Health metrics calculated 
by the [pat_dailySoH()](https://mazamascience.github.io/AirSensor/reference/pat_dailySoH.html) 
function. (`pat_dailySoHPlot()` runs `pat_dailySoH()` internally and uses
the output to create the plot.)

To better understand each plot, see also:

* [PurpleAirSoH_dailyPctReporting()](https://mazamascience.github.io/AirSensor/reference/PurpleAirSoH_dailyPctReporting.html), 
for `pm25_A_PctReporting` and `pm25_B_PctReporting`. Plots show the percentage of 
each day that the sensor is reporting data.

* [PurpleAirSoH_dailyPctValid()](https://mazamascience.github.io/AirSensor/reference/PurpleAirSoH_dailyPctValid.html), 
for `pm25_A_PctValid` and `pm25_B_PctValid`. Plots show percentage of the total 
recorded measurements that are considered physically plausible, _e.g._ no negative
values for pm2.5.

* [PurpleAirSoH_dailyPctDC()](https://mazamascience.github.io/AirSensor/reference/PurpleAirSoH_dailyPctDC.html), 
for `pm25_A_pctDC` and `pm25_B_pctDC`. A high percent DC (direct current) value 
indicates the likely occurrence of a “sticky value”, and a zero or low percent 
DC indicates that the sensor is recording dynamic data.

* [PurpleAirSoH_dailyMetFit()](https://mazamascience.github.io/AirSensor/reference/PurpleAirSoH_dailyMetFit.html), 
for `pm25_A_temperature_rsquared`. Plot shows daily linear model fit values between 
the pm25_A and temperature channels. The daily r-squared value is returned and is
expected to hover near 0 for a properly functioning sensor, under normal 
environmental conditions.

* [PurpleAirSoH_dailyABFit()](https://mazamascience.github.io/AirSensor/reference/PurpleAirSoH_dailyABFit.html), 
for `pm25_A_pm25_B_slope`, `pm25_A_pm25_B_intercept`, and `pm25_A_pm25_B_rsquared`.
Plots show daily linear model fit values between 
the pm25_A and pm25_B channels. The daily r-squared value is returned in addition to
the coefficients of the linear fit (slope and intercept).


```{r pat_dailySoHPlot-Liberty-School, eval = TRUE, warning = FALSE, message = FALSE}
# Create state-of-health plots for Liberty School
pat_dailySoHPlot(Liberty_School, ncol = 2)
```

The red line in each plot identifies expected values. 
The plot that grabs our attention the most is `pm25_A_temperature_rsquared` with 
the persistent elevated values occurring throughout the month. Values are expected to hover near 0 for 
a properly functioning sensor under normal environmental conditions. 
However, wildfire events during the summer and the heavy use of wood heat during winter
months could explain fluctuating r-squared values.
In our case, it's possible that the elevated values recorded in this plot are 
due to the wildfire events during that period.

The `pm25_A_pctDC` plot shows that channel B reported dynamic data throughout
the month while channel A recorded a higher percentage of "sticky values" 
towards the beginning and the end of the month. (Looking at the raw data, channel
A had periods where it reported repeated PM2.5 values of `0.0`.)
We notice that this overlaps with periods when the r-squared values were lower 
as shown in the `pm25_A_pm25_B_rsquared` plot. Especially at the beginning of the 
month, the plot shows r-squared values hovering around zero, 
indicating the poor functioning of the sensor during that period.

## pat_monitorComparison

The [pat_monitorComparison()](https://mazamascience.github.io/AirSensor/reference/pat_monitorComparison.html)
function is in another new category of functions that compare hourly aggregated sensor 
data with data from a professionally maintained "Federal Reference Method" (FRM) monitor

`pat_monitorComparison()` 
creates and returns a ggplot object that plots raw pat data, 
hourly aggregated pat data and hourly data from the nearest monitor 
from the US Forest Service maintained 
[PWFSL database](https://haze.airfire.org/monitoring/).

If the nearest monitor didn't collect data during the period of time investigated
or is not within the `distanceCutoff`, it is possible that you will encounter the following error: 

```
Error in monitor_subset(ws_monitor, tlim = c(starttime, endtime)) : ws_monitor object contains zero monitors
```

You can try to change the `distanceCutoff = 20` to 50 (km). However, 20 km is the 
recommended distance to ensure that both instruments are measuring similar parcels
of air.

To learn more info about the FRM monitors in your area of interest and 
discover their exact location, you can use the USFS AirFire 
[Monitoring Site](https://tools.airfire.org/monitoring/v4).

When aggregating raw data to hourly averages, `pat_monitorComparison()` applies
one of several Quality Control functions available in the **AirSensor**
package. These functions use various algorithms to invalidate certain hourly 
values that don't pass QC. The `FUN` argument lets you specify which QC algorithm to use. 

For specific details on the QC logic, have look at the online documentation available 
for each of the four algorithms: 

* [PurpleAirQC_hourly_AB_00](https://mazamascience.github.io/AirSensor/reference/PurpleAirQC_hourly_AB_00.html), 
* [PurpleAirQC_hourly_AB_01](https://mazamascience.github.io/AirSensor/reference/PurpleAirQC_hourly_AB_01.html), 
* [PurpleAirQC_hourly_AB_02](https://mazamascience.github.io/AirSensor/reference/PurpleAirQC_hourly_AB_02.html), 
* [PurpleAirQC_hourly_AB_03](https://mazamascience.github.io/AirSensor/reference/PurpleAirQC_hourly_AB_03.html).

The default "AB_01" algorithm attempts to find a balance between throwing out
data of interest and including data that is obviously junk. It works tolerably
well under normal conditions but is a little too strict during periods of very
high PM2.5 values like the Fourth of July and wildfire events.

```{r patmonitorComparison-Liberty-School, eval = TRUE, warning = FALSE, message = FALSE}
# comparison plot for Liberty School sensor
pat_monitorComparison(
  pat = Liberty_School,
  FUN = AirSensor::PurpleAirQC_hourly_AB_01, # default QC
  distanceCutoff = 20
)

# find out which is the nearest monitor
Liberty_School$meta$pwfsl_closestMonitorID # "530470010_01"
```

The graph shows:

* in gray -- sensor data before aggregation and QC (PA raw)
* in purple -- sensor data after the QC treatment (PA hourly)
* in black -- data collected by the nearest monitor (Monitor)

There is general agreement between "PA hourly" and "Monitor" but many of the 
hourly aggregated values are missing because they have been invalidated by our
QC function.

Because we have looked carefully at the raw data from Liberty School and know 
this sensor is functioning well, we feel comfortable use the very forgiving "AB_00"
QC function to see more of the hourly aggregated values:

```{r patmonitorComparison-Liberty-School-QC00, eval = TRUE, warning = FALSE, message = FALSE}
# comparison plot for Liberty School sensor
pat_monitorComparison(
  pat = Liberty_School,
  FUN = AirSensor::PurpleAirQC_hourly_AB_00, # minimal QC
  distanceCutoff = 20
)
```

We observe some agreement between the PA hourly and the monitor data. The trends
are similar but the absolute values differ. A more detailed assessment of their 
statistical relationship is discussed in the next section. 
 
## pat_externalFit

The [pat_externalFit()](https://mazamascience.github.io/AirSensor/reference/pat_externalFit.html)
function produces a linear model between data from PurpleAir and data from 
the closest FRM monitor. A diagnostic plot is produced if 
`showPlot = TRUE`.

_NOTE:_ As of **AirSensor** version 1.0.8, this function does not properly
support choosing the QC algorithm and `hourly_AB_01` is always used.

```{r patexternalFit-01-Liberty-School, eval = TRUE, warning = FALSE, message = FALSE}
# create comparison plot for Liberty School sensor
lm_01 <- pat_externalFit(
  pat = Liberty_School,
  showPlot = TRUE
)

summary(lm_01)
```

We can _almost_ see that the nearest FRM monitor to the Liberty School sensor
is 4.2 km away -- nicely close for this analysis.

Unfortunately, several of the hourly aggregated values have been invalidated 
during the periods of highest wildfire smoke. Of the remaining hourly values, the
linear fit is actually pretty good with an r-squared of 0.94 and a slope of 0.69
implying that the PurpleAir sensor is overstating the PM2.5 values.

Overall, the Liberty sensor tracked wildfire smoke intrusions quite well.

----

_Best of luck assessing air quality in your community!_
