---
title: "Australia on Fire"
author: "Kayleigh Wilson"
date: "2020-01-05"
output:
  html_document:
    toc: true
    theme: united
---

```{r setup, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 7, fig.height = 5)
```

## Goals

The goal of this report is to demonstrate the use of the **AirSensor** package
to create a local archive or PurpleAir data covering the December 2019 - 
January 2020 fires and smoke in Australia.

The target audience is any R user interested in air quality data but who who may 
not be familiar with Mazama Science pacakges. 

We begin by loading the necessary packages:

```{r load_packages, message = FALSE}
library(dplyr)            # data manipulation
library(ggplot2)          # plotting
library(PWFSLSmoke)       # USFS monitor data access and plotting
library(AirSensor)        # PurpleAir sensor data access and plotting
library(AirMonitorPlots)  # Advanced plots for monitors
```

# *TODO: Explain the difference between 'monitors' and 'sensors'.*

Data can come from two different sources, either a 'monitor' or a 'sensor'. 
Within the AirSensor package, we consider a 'monitor' to be something installed
and operated federally, in contrast to a 'sensor' which is a PurpleAir sensor, 
installed and operated by anyone. 

## Getting the Data

### Set up a local archive

*TODO: Explain what this code does.*

First, if the archive base directory does not exist, we can set it up so that 
RStudio knows where to find and write data.  


```{r setup_local_archive}
archiveBaseDir <- path.expand("~/Data/Australia_on_fire")
if ( !dir.exists(archiveBaseDir) ) {
  dir.create(archiveBaseDir)
}
setArchiveBaseDir(archiveBaseDir)
```

### Load synoptic data for Australia

*TODO: Explain what this code does.*

Next, make sure the Australia pas exists in your directory. The following chunk 
of code will check to to see if it exists, if not, it will download it.


```{r load_synoptic}
filePath <- file.path(archiveBaseDir, "pas_au.rda")

if ( !file.exists(filePath) ) {
  initializeMazamaSpatialUtils()
  pas_au <- pas_createNew(countryCodes = "AU", includePWFSL = TRUE)
  save(pas_au, file = filePath)
}

pas <- get(load(filePath))
```

### View the 'pas' data

*TODO: Explain what this code does.*

We can view the locations of each sensor and the AQI over the last hour using the 
pas_leaflet function.

```{r view_pas}
pas_leaflet(pas)
```

### Load time series data for Australia

It looks like the air quality in Sydney is currently unhealthy, let's look at
data from a sensor there. 

*TODO: Create code similar to the the "load_synoptic" code above to install/load
2-8 weeks of pat data for some/all of the sensors in Australia. It's OK if this
takes a few minutes the first time.*

```{r load_pat}
label <- c("Hurstville")

deviceDeploymentID <- pas_getDeviceDeploymentIDs(pas = pas, 
                                                 countryCodes = c("AU"), 
                                                 pattern = label)

pat_hurstville <- pat_createNew(pas = pas, id = deviceDeploymentID, 
                          startdate = 20191201, enddate = 20200108)

```


### View the 'pat' data

*TODO: Simple example looking at a sensor in a major population center.*

```{r view_pats}

pat_multiplot(pat_hurstville)

```


## Explore the 'pas' dataframe

A 'pas' (PurpleAir synoptic) object is a dataframe that contains metadata and 
PM 2.5 averages for many purple air sensors in a designated region. Each pas 
object can be filtered and edited to contain whichever arrangement of sensors 
that the programmer desires based on location, state, name, etc, due to the 
flexibility of being a dataframe. Data averages that the pas object contain 
are current at the time that the pas is created and are different from time 
series data which are reserved for the 'pat' objects. 

For the purposes of this exploratory example, we are focusing on Australia but 
maybe we want to filter even more and just look at the sensors within a certain 
radius of the one we chose in Sydney. 

*TODO: Describe the pas object and show how to maipulate and visualizae data
using other `pas_~() functions. There could be subsections in here.*

```{r explore_pas}
lon <- pat_hurstville$meta$longitude
lat <- pat_hurstville$meta$latitude

pas_sydney <- 
  pas_au %>%
  pas_filterNear(longitude = lon, latitude = lat, radius = "60 km")

pas_leaflet(pas_sydney)

```


*TODO: You may choose to only show the code for some examples by running a
chunk with `eval = FALSE`.*

## Explore the 'pat' data

A 'pat' (PurpleAir timeseries) object is a list of two dataframes, one metadata 
dataframe, and one timeseries dataframe for one sinlge PurpleAir sensor. For the
purposes of this example, 



*TODO: Describe the pat object and show how to manipulate and visualize data 
using other `pat_~() functions. There will be subsections in here.*



## Working with 'airsensor' data

*TODO: Describe creation of 'airsensor' objects and how the __PWFSLSmoke__
package can be used to manipulate and visualize them. There will be subsections 
in here.*

## Advanced plotting with **AirMonitorPlots**

*TODO: Show a couple of examples.*

## STORY

*TODO:  Pick one of sensors in a major population center and work up a story.*

