---
title: "Australia on Fire"
author: "Kayleigh Wilson & Jonathan Callahan"
date: "2020-01-15"
output:
  html_document:
    toc: true
    theme: united
---

```{r setup, include=FALSE, message = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  fig.width = 7, 
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
```

## Goals

The goal of this report is to demonstrate the use of the **AirSensor** package
to explore the smoke impacts of the December 2019 - January 2020 fires and smoke
in Australia. This document demonstrates "reproducible research" in that no
pre-generated files are needed. Raw data from PurpleAir will be downloaded
and processed when this script is first run.

We begin by loading the necessary packages:

```{r load_packages, message = FALSE}
library(dplyr)            # data manipulation
library(ggplot2)          # plotting
library(PWFSLSmoke)       # USFS monitor data access and plotting
library(AirSensor)        # PurpleAir sensor data access and plotting
library(AirMonitorPlots)  # Advanced plots for monitors
```

In this document we will use the term "sensor" to refer to consumer-grade,
PurpleAir air quality sensors sampling at a high time resolution. The term 
"monitor" will refer to professional-grade, typically government managed 
monitors with hourly data.

The **AirSensor** package has three data models:

* _pas_ (**P**urple**A**ir **S**ynoptic) objects contain snapshots of multiple PA 
sensors and can be used to make maps
* _pat_ (**P**urple**A**ir **T**imeseries) objects contain high resolution data for a 
single sensor
* _airsensor_ objects contain PA timeseries data that has been quality controlled
and aggregated to an hourly time axis.

## Getting the Data

### Set up a local archive

The **AirSensor** package needs to know where processed data will live. For this 
report, we will specify a local `archiveBaseDir` where downloaded and processed
data will live. The following code specifies a local directory archive, checks 
for its existence and creates it if needed.

```{r setup_local_archive}
archiveBaseDir <- path.expand("~/Data/Australia_on_fire")
if ( !dir.exists(archiveBaseDir) ) {
  dir.create(archiveBaseDir)
}

setArchiveBaseDir(archiveBaseDir)
```

### Load synoptic data for Australia

We will use the `pas_createNew()` function to create a _pas_ object containing
all the spatial metadata associated with purple air monitors in Australia.

The following code checks to make sure the `pas_au` file exists in your 
`archiveBaseDir` directory and, if missing, recreates it.

```{r load_synoptic}
filePath_pas <- file.path(archiveBaseDir, "pas_au.rda")

if ( !file.exists(filePath_pas) ) {
  initializeMazamaSpatialUtils()
  pas_au <- pas_createNew(countryCodes = "AU", includePWFSL = TRUE)
  save(pas_au, file = filePath_pas)
}

pas <- get(load(filePath_pas))
```

### View a _pas_ map

A _pas_ object is a dataframe that contains metadata and 
PM 2.5 averages for many purple air sensors in a designated region. Each pas 
object can be filtered and edited to retain whichever collection of sensors 
the analyist desires based on location, state, name, _etc._

It is important to note that the data averages in the _pas_ object -- the numeric
values for PM2.5 or temperature or humidity -- are current at the time that the 
pas is created. _pas_ objects can be used to quickly explore the spatial 
distribution of PurpleAir sensors and display some _then_-current values but 
should not be used for detailed analysis.

We can view the locations of each sensor and the AQI (Air Quality Index) maxima
(when the _pas_ was created) using the `pas_leaflet()` function.

```{r view_pas}
pas_leaflet(pas)
```

### Load time series data for a sensor

On January 15, 2020, the leaflet plot showed unhealthy air quality for Sydney. 
Let's look at data from a sensor in the Sydney area for the time period covering
the Australian wildfires. One of those sensors is labeled "Jannali"

The following chunk will check to see if the _pat_ object for the sensor named
"Jannali" has already been created the local directory and will recreate it if
needed.

```{r load_pat}
label <- c("Jannali" )

filePath_jannali <- file.path(archiveBaseDir, "pat_jannali.rda")

if ( file.exists(filePath_jannali) ) {
  pat_jannali <- get(load(filePath_jannali))
} else {
  pat_jannali <- pat_createNew(
    pas = pas, 
    label = label, 
    startdate = 20191229, 
    enddate = 20200110
  )
  save(pat_jannali, file = filePath_jannali)
}
```

### View a _pat_ timeseries

A _pat_ object is a list of two dataframes, one called
`meta` containing spatial metadata associated with the sensor and another called
`data` containing that sensor's time series data. Each _pat_ object contains
time series data for a temperature channel, a humidity channel, 
A and B PM 2.5 channel's, and several other fields. 

The following chunk demonstrates use of the `pat_multiPlot()` function to have a 
quick look at the data contained in a _pat_ object. The plot shows both A and B 
channels as well as temperature and humidity. The plotting function is flexible 
and has options for choosing which channels to display, on the same or 
individual axes. (_Type `?pat_multiPlot` to learn more._)

In this case we see that the data look good. Channels A and B are both showing
lots of PM 2.5 data for "Jannali" over the period of interest 

```{r view_pats}
pat_multiPlot(pat_jannali)
```

## Explore the _pas_ object

For the purposes of this exploratory example, we are focusing on Australia but 
maybe we want to filter even more and just look at the sensors within a certain 
radius of the one we chose in Sydney. 


```{r explore_pas}
lon <- pat_jannali$meta$longitude
lat <- pat_jannali$meta$latitude

pas_sydney <- 
  pas_au %>%
  pas_filterNear(
    longitude = lon, 
    latitude = lat, 
    radius = "50 km"
  ) 

pas_leaflet(pas_sydney)
```

### Filter the _pas_ based on conditions

On Januar 15, 2020, the air quality in Sydney appeared to be unhealthy. 
[_Using a _pas_ object created don that day_] we can look for the 
sensors that are reporting the highest `pm25` levels.

The next chunk of code filters 
the _pas_ for the sensors that have a reported `pm25` value higher than 150. 

It is important to note here that the sensors listed in the _pas_ after 
filtering will depend on the date that the original _pas_ was generated. Since 
data values are only valid for the time a _pas_ is created, the _pas_ that you use
may show different for the same sensors generated with _pas_ that we are 
using here. This document was created using a _pas_ with data from January 
15, 2020.

```{r unhealthy_pas}
pas_high_pm25 <- 
  pas_au %>%
  pas_filter(pm25_1hr > 150)

if ( nrow(pas_high_pm25) > 1 ) {
  pas_leaflet(pas_high_pm25)  
}
```

## Explore the _pat_ data

### Download _pat_ data

Now that the _pas_ has been filtered based on the worst air quality, we can 
create _pat_ objets for sensors listed in the _pas_. You could 
download all the sensors in the newly filtered _pas_ but, for the
purposes of reproducibility, we will continue to use the `pas_au` object 
as the reference _pas_ to ensure that the sensors referenced are present. 
Again, this is due to the ephemeral nature of the data in a _pas_ object which 
is only valid when it is initially created..

Since the fires in Australia started, new PurpleAir sensors have been popping up 
left and right. Let's start by grabbing data from sensors with the longest 
history. The following chunk will check to see if the _pat_ objects are already
in the local data directory, if not, they will be downloaded.

```{r explore_pat}

startdate <- 20191210
enddate <- 20200110

filePath_chisholm <- file.path(archiveBaseDir, "pat_chisholm.rda")
if ( file.exists(filePath_chisholm) ) {
  pat_chisholm <- get(load(filePath_chisholm))
} else {
  pat_chisholm <- pat_createNew(
    label = "Chisholm",
    pas = pas_au,
    startdate = startdate,
    enddate = enddate
  )
}

filePath_moruya <- file.path(archiveBaseDir, "pat_moruya.rda")
if ( file.exists(filePath_moruya) ) {
  pat_moruya <- get(load(filePath_moruya))
} else {
  pat_moruya <- pat_createNew(
    label = "MORUYA HEADS",
    pas = pas_au,
    startdate = startdate,
    enddate = enddate
  )
} 

filePath_windang <- file.path(archiveBaseDir, "pat_windang.rda")
if ( file.exists(filePath_windang) ) {
  pat_windang <- get(load(filePath_windang))
} else {
  pat_windang <- pat_createNew(
    label = "Windang, Ocean Street",
    pas = pas_au,
    startdate = startdate,
    enddate = enddate
  )
} 

```

In order to look for patterns, we can look at the PM2.5 data recorded on channel
A from all the sensors. This chunk uses **ggplot2** to view all the data on the same
axis. 

```{r plot_pats, message=FALSE, warning=FALSE}
colors <- c(
  "Chisholm" = "#1b9e77", 
  "Moruya" = "#d95f02", 
  "Windang" = "#7570b3"
)

gg <- ggplot(data = pat_chisholm$data) +
  geom_point(aes(x = pat_chisholm$data$datetime, 
                 y = pat_chisholm$data$pm25_A, 
                 color = "Chisholm"), alpha = 0.5) +
  geom_point(data = pat_moruya$data, 
             aes(x = pat_moruya$data$datetime, 
                 y = pat_moruya$data$pm25_A,
                 color = "Moruya"), alpha = 0.5) +
  geom_point(data = pat_windang$data, 
             aes(x = pat_windang$data$datetime, 
                 y = pat_windang$data$pm25_A, 
                 color = "Windang"), alpha = 0.5) +
  labs(title = "PM 2.5 channel A for multiple sensors" ) +
  xlab("date") +
  ylab("ug/m3") +
  scale_colour_manual(name="Sensor",values=colors) +
  theme(legend.position= c(0.9, 0.8))

print(gg)
```

Values above 1000 μg/m³ may be suspect as they are above the maximum range
of the PLANTOWER particle detectors used in Purple Air sensors.
(See [PLANTOWER PMS5003 specs](http://www.aqmd.gov/docs/default-source/aq-spec/resources-page/plantower-pms5003-manual_v2-3.pdf))

### Create _pat_ objects based on proximity

Let's check a few other sensors that are closer in proximity to Chisholm to 
see if they are also reporting such high values.

```{r proximity_pats, message=FALSE, warning=FALSE}
filePath_bungendore <- file.path(archiveBaseDir, "pat_bungendore.rda")
if ( file.exists(filePath_bungendore) ) {
  pat_bungendore <- get(load(filePath_bungendore))
} else {
  pat_bungendore <- pat_createNew(
    label = "Bungendore, NSW Australia", 
    pas = pas_au,
    startdate = 20191229, 
    enddate = 20200110
  )
  save(pat_bungendore, file = filePath_bungendore)
}

filePath_downer <- file.path(archiveBaseDir, "pat_downer.rda")
if ( file.exists(filePath_downer) ) {
  pat_downer <- get(load(filePath_downer))
} else {
  pat_downer <- pat_createNew(
    label = "Downer",
    pas = pas_au,
    startdate = 20191220, 
    enddate = 20200110
  )
  save(pat_downer, file = filePath_downer)
}
```

These recently installed sensors have a shorter data history but are located
close to our Sydney sensors and we would expect the PM2.5 signals to be similar. 

The next chunk uses **ggplot2** to plot A channel
data from the "Chisholm ACT Australia" sensor in the same green color as  
above in addition to channel A data from the two nearest sensors.


```{r proximity_plot, message=FALSE, warning=FALSE}
colors <- c("Chisholm" = "#1b9e77", 
            "Downer" = "#f1a340", 
            "Bungendore" = "#998ec3")

gg <- ggplot(data = pat_chisholm$data) +
  geom_point(aes(x = pat_chisholm$data$datetime, 
                 y = pat_chisholm$data$pm25_A, 
                 color = "Chisholm"), alpha = 0.5) +
  geom_point(data = pat_downer$data, 
             aes(x = pat_downer$data$datetime, 
                 y = pat_downer$data$pm25_A,
                 color = "Downer"), alpha = 0.5) +
  geom_point(data = pat_bungendore$data, 
             aes(x = pat_bungendore$data$datetime, 
                 y = pat_bungendore$data$pm25_A, 
                 color = "Bungendore"), alpha = 0.5) +
  labs(title = "PM 2.5 channel A for multiple sensors" ) +
  xlab("date") +
  ylab("ug/m3") +
  scale_colour_manual(name = "Sensor", values = colors) +
  theme(legend.position = c(0.9, 0.8))

print(gg)
```

All three sensors are reporting very high values on the same 
days between December 30, 2019 and January 6, 2020. We can use **AirSensor**
package "State-of-Health" functions to evaluate whether these recently installed
sensors are performing properly.

### Sensor State-of-Health

After looking at sensor data from numerous sensors over long periods of time, we
have noticed that the _pat_ data quality can degrade over time. A quick way to 
quantify data quality is to simply calculate the r^2^ value between channels A 
the daily data quality is categorized as good, fair, or poor. Typically, new 
sensors have not had time for outside factors to degrade data quality, but for a 
quick sanity check, we can use the `pat_dailySoHIndexPlot()` function to plot the
daily State-of-Health index. This function plots both channels A and B with 
a daily State-of-Health index along the bottom.

The State-of-Health index is a multi-metric index and with several components
including. The `dailyPctReporting` component is most likely responsible for the
brief segment of "Poor" during the first partial day in the time series.

```{r pat_soh, message=FALSE, warning=FALSE}
pat_dailySoHIndexPlot(pat_chisholm)
```

The sensor seems to be in good health so. It is uncler what to make of such
high PM2.5 values and whether they are out-of-spec or not. In any case, the
smoke in the Sydney area was clearly very bad over the time period of interest.

### Sensor A/B agreement

The primary component used in the State-of-Health index is the level of agreement
between the A and B channel measurements. The `pat_internalFit()` function makes it easy to
compare these values and calculate the R^2^ .

```{r pat_internalFit}
pat_internalFit(pat_chisholm)
```

With an R^2^ of 0.998, this sensor seems to be performing quite well indeed.

## Working with _airsensor_ data

Another way to look at the PurpleAir sensor data is to convert the _pat_ into an
_airsensor_ object. The following chunk aggregates data from a pat object into an 
_airsensor_ object with an hourly time axis. This new _airsensor_ object is
equivalent to a _ws_monitor_ object used in the **PWFSLSmoke** package.

Because we are dealing with values far outside the norm, we will use the
`PurpleAirQC_hourly_AB_00` function which performs minimal QC.

````{r airsensor_data}
airsensor_chisholm <- pat_createAirSensor(
  pat = pat_chisholm,
  parameter = "pm25",
  FUN = PurpleAirQC_hourly_AB_00
)
```

### Advanced plotting

Now that we have converted the relative raw _pas_ into an _airsensor_ object, 
we can any of the "monitor" plotting functions found in the **PWFSLSmoke** or
*AirMonitorPlots* packages

```{r monitor_plots}
AirMonitorPlots::monitor_ggDailyBarplot(airsensor_chisholm)
```

Here, the bar associated with each day is colored by Air Quality Index (AQI). 
Over this time  period there were only 10 days where the daily average AQI was
below "Unhealthy". Yikes!

Unfortunately, the current QC algorithm in the package invalidates values over
1000 resulting in missing daily averages on the days with the highest measurements.
(_This threshold is being reevaluated._)

### Incorporating wind data

To get a sense of what direction smoke is coming from, we use the  
`sensor_PolluationRose()` function. As the name implies, this function takes
an _airsensor_ objecct as an argument. It then obtains hourly 
wind direction and speed data from the nearest meterological site and plots a 
traditional wind rose plot for wind direction and PM2.5.

In this case, it looks like the smoke is coming mostly from the E/NE which is 
validated by this wind rose plot from the Australian Bureau of Meteorology here:
http://www.bom.gov.au/cgi-bin/climate/cgi_bin_scripts/windrose_selector.cgi?period=Dec&type=9&location=72150

```{r pollution_rose}
sensor_pollutionRose(sensor = airsensor_chisholm)
```

The `sensor_pollutionRose()` function uses the **metdata** package to obtain
wind data. We can use this package ourselves to get location information for
the narest meteorological site to Chisholm so that we can map the location
relative to our PurpleAir sensors.

The chunk below finds the meteorological site closest to Chisholm, and plots it 
on the map in blue along with oter sensors near Canberra.

```{r leaflet_map}
# Target location
lon <- pat_chisholm$meta$longitude
lat <- pat_chisholm$meta$latitude

# Meteorological sites
closestSite <- worldmet::getMeta(lon = lon, 
                                 lat = lat, 
                                 n = 1, 
                                 plot = FALSE)[1,]

# Canberra sensors
pas_canberra <- 
  pas_au %>%
  pas_filterNear(
    latitude = pat_chisholm$meta$latitude, 
    longitude = pat_chisholm$meta$longitude, 
    radius = "40 km"
  )

# Leaflet map
map <- pas_leaflet(pas_canberra) 

map <-
  map %>%
  leaflet::addCircleMarkers(lng = closestSite$longitude,
                            lat = closestSite$latitude)

map
```

Hopefully, this tour of functionality gives you a sense of the capabilities of
the **AirSensor** package and provides a template for your own investigations of
air quality data.

_Happy Exploring!_

_Mazama Science_























