---
title: "Creating Data Archives"
author: "Jonathan Callahan"
date: "2023-04-03"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating Data Archives}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 5)
```

## Introduction

This tutorial demonstrates how to create a multi-month archive of PurpleAir
timeseries data that can be loaded from a local `archiveBaseDir`. Target
audiences include grad students, researchers and any member of the public
concerned about air quality and comfortable working with R and RStudio.

## Goal

The goal in this tutorial is to explain the file naming protocol
and directory structure required for data to be loaded using the `pas_load()`,
pat_load()` and `sensor_load()` functions. We will build a multi-month archive 
for SCAQMD and demonstrate accessing that data.

## Directory Structure

The *AirSensor* package has a variety of "load" functions, all of which assume
that pre-generated data files will be found in specific locations underneath
an archive directory. The base of this directory is specified by 
`setArchiveBaseUrl()` if the archive is web based or `setArchiveBaseDir()` if
the archive is local.

For a multi-year archive, the overall structure beneath the `archiveBaseDir` 
will look something like this, depending on how many years of data you have:

```
├── airsensor
│   ├── 2022
│   ├── 2023
│   └── latest
├── pas
│   ├── 2022
│   └── 2023
├── pat
│   ├── 2022
│   │   ├── 01
│   │   ├── ...
│   │   └── 12
│   ├── 2023
│   │   ├── 01
│   │   ├── 02
│   │   └── 03
│   └── latest
```

## File Naming Protocol

The protocol for creating pre-generatededata files depends on the type of object
being created.

### airsensor

* Directory: `<archiveBaseDir>/airsensor/<YYYY>/`
* File name: `airsensor_<collectionName>_<YYYYMM>.rda`
* Example: `~/Data/SCAQMD/airsensor/2023/airsensor_scaqmd_202304.rda`

_NOTE:_ `<collectionName>` is an arbitrary character string defined by the user.  

### pas

* Directory: `<archiveBaseDir>/pas/<YYYY>/`
* File name: `pas_<YYYYMMDD>.rda`
* Example: `~/Data/SCAQMD/pas/2023/pas_20230403.rda`

### pat

* Directory: `<archiveBaseDir>/pas/<YYYY>/<MM>/`
* File name: `pat_<deviceDeploymentID>_<YYYYMM>.rda`
* Example: `~/Data/SCAQMD/pat/2023/04/pat_ab5dca99422f2c0d_13669_202304.rda`

## _load() Functions

When loading data, package `~_load()` functions look for any local archive 
specified with `setArchiveBaseDir()`. If no local archive is found, a web based
archive is searched at the URL specified with `setArchiveBaseUrl()`.

Package functions that load pre-generated data files include:

* `pas_load()`
* `pat_load()`
* `pat_loadLatest()`
* `pat_loadMonth()`
* `sensor_load()`
* `sensor_loadLatest()`
* `sensor_loadMonth()`

To avoid internet latency, specification of `archiveBaseDir` will always take precedence 
over specification of `archiveBaseUrl`. For this reason, if you previously set `archiveBaseDir`
and you then want to load data from some `archiveBaseUrl`, you will have to 
first: `setArchiveBaseDir(NULL)`.

## R script

The following R script will take several minutes to run and will create an
archive of 'pat' and 'airsensor' data files on your computer. Once these 
files have been created, loading them will be very fast.

After running the script, a final 
section will demonstrate how to load and work with these local data files.

This R script can be used as a starting point for anyone interested in creating 
small collections of data for other communities and other dates.

```{r create-archive, eval = FALSE, warning = FALSE, message = FALSE}
# SCAQMD local data archive: Setup

library(sf)                 # Helper for spatial data
library(MazamaCoreUtils)    # Logging and time handling
library(MazamaSpatialUtils) # Spatial metadata assignment
library(AirSensor)

# Set API keys
source("./global_vars.R") # See "Working with PurpleAir API Keys"

# ----- Setup ------------------------------------------------------------------

# Have log messages sent to the console
MazamaCoreUtils::logger.setLevel(TRACE)

# Use the default archiveDir unless it is already defined
if ( !exists("archiveDir") ) {
  archiveDir <- file.path("~/Data/SCAQMD")
}

# Set the package archiveBaseDir so we can load pat objects with `pat_load()`
setArchiveBaseDir(archiveDir)

# Create datestamps
timezone <- "UTC"
datestamp <- strftime(lubridate::now(tzone = timezone), "%Y%m%d", tz = timezone)
monthstamp <- stringr::str_sub(datestamp, 1, 6)
yearstamp <- stringr::str_sub(datestamp, 1, 4)

# ----- Create PAS object ------------------------------------------------------

# Set up MazamaSpatialUtils
AirSensor::initializeMazamaSpatialUtils()

# Create the archiveDir/pas/YYYY directory
dir.create(
  file.path(archiveDir, "pas", yearstamp), 
  showWarnings = FALSE,
  recursive = TRUE
)

pas <-
  pas_createNew(
    api_key = PURPLE_AIR_API_READ_KEY,
    countryCode = "US",
    stateCode = "CA",
    show_only = "143602,22727,23241",       # Change to your list of sensor indices
    lookbackDays = 1,
    location_type = 0
  )

# Save the "latest" version
filename <- paste0("scaqmd_", datestamp, ".rda")
filepath <- file.path(archiveDir, "pas", yearstamp, filename)
logger.info("Writing PAS data to %s", filename)
save(list = "pas", file = filepath)

# ----- Prepare PAT info -------------------------------------------------------

# Get the sensor index values
scaqmd_ids <- pas_getDeviceDeploymentIDs(pas)

# Set up months
timezone <- "America/Los_Angeles"
monthStamps <- c(202301, 202302, 202303)

# Loop over months
for ( monthStamp in monthStamps ) {
  
  logger.debug("Working on monthStamp %s ---------- ...", monthStamp) 
  
  # Get POSXct startdate
  startdate <- MazamaCoreUtils::parseDatetime(monthStamp, timezone = timezone)
  
  # Guarantee that the enddate is the first of the next month
  enddate <- lubridate::floor_date(
    startdate + lubridate::ddays(40),
    unit = "month"
  )
  
  # Get YYYY and MM strings
  YYYY <- strftime(startdate, "%Y")
  MM <- strftime(startdate, "%m")
  
  # Initialize counters
  idCount <- length(scaqmd_ids)
  count <- 0 
  successCount <- 0
  
  # ----- Create PAT objects ---------------------------------------------------
  
  # Create the archiveDir/pat/YYYY/MM/ directory
  dir.create(
    file.path(archiveDir, "pat", YYYY, MM), 
    showWarnings = FALSE,
    recursive = TRUE
  )
  
  # Loop over all deviceDeploymentIDs
  for ( id in scaqmd_ids ) {
    
    sensor_index <- 
      pas %>% 
      dplyr::filter(deviceDeploymentID == id) %>%
      dplyr::pull(sensor_index)
    
    # Create PAT canonical file name
    fileName <- paste0("pat_", id, "_", YYYY, MM, ".rda")
    
    # Create PAT canonical file path
    filePath <- file.path(archiveDir, "pat", YYYY, MM, fileName)
    
    count <- count + 1
    logger.debug("Working on %s (%d/%d) ...", id, count, idCount)
    
    # Use a try-block in case you get "no data" errors
    result <- try({
      
      # Create PAT
      pat <- pat_createNew(
        api_key = PURPLE_AIR_API_READ_KEY,
        pas = pas,
        sensor_index = sensor_index,
        startdate = startdate,
        enddate = enddate,
        timezone = "UTC",
        average = 0,
        verbose = FALSE
      )
      
      successCount <- successCount + 1
      save(pat, file = filePath)
      
    }, silent = FALSE)
    
    if ( "try-error" %in% class(result) ) {
      logger.error(geterrmessage())
    }
    
  }
  
  # ------ Create AirSensor objects --------------------------------------------
  
  # Create the archiveDir/airsensor/YYYY/ directory
  dir.create(
    file.path(archiveDir, "airsensor", YYYY), 
    showWarnings = FALSE,
    recursive = TRUE
  )
  
  # Assign a collection name that makes sense
  collectionName <- "scaqmd"
  
  # Init counts
  successCount <- 0
  count <- 0
  
  dataList <- list()
  
  # Loop over all ids and aggregate to hourly
  for ( id in scaqmd_ids ) {
    
    count <- count + 1
    
    # Debug info
    logger.debug(
      "%4d/%d Calling pat_createAirSensor('%s')",
      count,
      length(scaqmd_ids),
      id
    )
    
    # Load the pat data, convert to an airsensor and add to dataList
    dataList[[id]] <- tryCatch(
      expr = {
        airsensor <- pat_load(
          id = id,
          label = NULL,
          pas = scaqmd,
          startdate = startdate,
          enddate = enddate,
          timezone = "America/Los_Angeles"
        ) %>%
          pat_createAirSensor(
            FUN = AirSensor::PurpleAirQC_hourly_AB_01
          )
      }, 
      error = function(e) {
        logger.warn('Unable to load PAT data for %s ', id)
        NULL
      }
      
      # Keep going in the face of errors
    )
    
  } # END of deviceDeploymentIDs loop
  
  # Combine the airsensors into a single airsensor object and save
  tryCatch(
    expr = {
      logger.info('Combining airsensors...')
      
      airsensor <- PWFSLSmoke::monitor_combine(dataList)
      class(airsensor) <- c("airsensor", "ws_monitor", "list")
      
      logger.info('Combined successfully...')
      
      # Create Airsensor canonical file name
      fileName <- paste0("airsensor_", collectionName, "_", YYYY, MM, ".rda")
      
      # Create Airsensor canonical file path
      filePath <- file.path(archiveDir, "airsensor", YYYY, fileName)
      
      save(list = "airsensor", file = filePath)
    }, 
    error = function(e) {
      msg <- paste("Error creating monthly AirSensor file: ", e)
      logger.error(msg)
    }
  )
  
  # Now proceed to the next month
}
```

## Loading Local Data

Our local archive of data was created following the directory and file
naming protocol. We can now work with this archive by running
`setArchiveBaseDir()`.

```{r load-local-data, eval = FALSE, warning = FALSE, message = FALSE}
library(AirSensor)
library(AirMonitorPlots)

# Use the tutorial default archiveDir unless it is already defined
if ( !exists("archiveDir") ) {
  archiveDir <- file.path("~/Data/SCAQMD")
}

# Set your archive base directory and check that is correct
setArchiveBaseDir(archiveDir)
getArchiveBaseDir()

# Load the custom pas file directly
pas <- pas_loadLatest()

# Use the interactive map to quickly get the "deviceDeploymentID" of the sensors
# you want to explore further based on their location. 
pas_leaflet(pas)

# However, data for a specific sensor might not be available for a given month.
# You can make sure to have the expected pat file in your month directory by 
# looking for that file.
list.files(file.path(archiveDir, "pat/2023/04"))

# Load specific days of pat data for Pine Forest
Pine_Forest  <- 
  pat_load(
    id = "0cbfeb2ce4c1553c_13661",
    pas = pas, 
    startdate = 20230201,
    enddate = 20230208, 
    timezone = "America/Los_Angeles"
  )

# Basic plot for Pine_Forest 
pat_multiplot(Pine_Forest)

# Load the all-hourly-data-combined sensor object
all_sensors <- 
  sensor_load(
    collection = "scaqmd", # collectionName defined by user 
    startdate = 20230201,
    enddate = 20230308,
    timezone = "America/Los_Angeles"
  )

# Plot all hourly values with daily averages
all_sensors %>%
  AirMonitorPlots::ggplot_pm25Timeseries() + 
  ggplot2::ggtitle("SCAQMD Smoke -- March, 2023") +
  AirMonitorPlots::geom_pm25Points(shape = "square", alpha = .1) + 
  AirMonitorPlots::stat_dailyAQCategory(alpha = .5) + 
  ggplot2::scale_y_continuous(limits = c(0, 300)) +
  AirMonitorPlots::custom_aqiStackedBar(width = 0.01) 
```

----

_Best of luck assessing air quality in your community!_
