---
title: "Custom QC Algorithms"
author: "Mazama Science"
date: "2020-06-19"
output: html_document
---

Data quality-control is the process of flagging or removing invalid or noise 
data points from a data set, either manually or algorithmically. There are
many statistical techniques and methods to preform quality-control. In this 
article we provide a general overview of how to construct such algorithms 
for use within the **AirSensor** package. Our goal is to write custom 
quality-control functions that replace, flag, or remove invalid data points.
Data that are out of an instrument's specifications, that appear to be 
associated with random noise or are associated with other known issues can be 
considered invalid.

NOTE: If you haven't yet read the _Temporal Aggregation_ article, you should
read it first in order to understand advanced usage of `pat_aggregate()`.

## Example 1 -- Count based QC

Let's start by creating a simple QC algorithm based on the count of the
number of data records within an hour. If the count of records is too low, we 
will replace any aggregated value with `NA`, otherwise we will return the 
hourly aggregated value, in this example the `mean`.

This QC function will create two hourly-aggregated 'pat' objects -- `mean_pat` 
and `count_pat` and use data from `count_pat` to invalidate some records in
`mean_pat`.


```{r}
library(AirSensor)
library(dplyr)

# Load a PurpleAir Timeseries ('pat') object
pat <- example_pat

# Define a QC function that returns an aggregated 'pat' object
QC_hourly_count <- function(pat, minCount = 11) {

  # Create mean hourly aggregate pat
  mean_pat <- pat_aggregate(pat, function(x) mean(x, na.rm = TRUE))

  # Create count of non-missing pat data
  count_pat <- pat_aggregate(pat, function(x) length(na.omit(x)))

  # Create a minimum count mask
  pm25_A_mask <- count_pat$data$pm25_A < minCount
  pm25_B_mask <- count_pat$data$pm25_B < minCount

  ##################
  # dplyr/tidyverse version
  ##################
  #pat <- mean_pat
  # mean_pat$data <-
  #   mean_pat$data %>%
  #   mutate(pm25_A = replace(.data$pm25_A, pm25_A_mask, NA)) %>%
  #   mutate(pm25_B = replace(.data$pm25_B, pm25_B_mask, NA)) %>%
  #   mutate(hr_count_pm25_A = count_pat$data$pm25_A) %>%
  #   mutate(hr_count_pm25_B = count_pat$data$pm25_B)
  # pat <- mean_pat

  ##################
  # base R version
  ##################
  pat <- mean_pat
  pat$data$pm25_A[pm25_A_mask] <- NA
  pat$data$pm25_B[pm25_B_mask] <- NA
  pat$data$hr_count_pm25_A <- count_pat$data$pm25_A
  pat$data$hr_count_pm25_B <- count_pat$data$pm25_B

  return(pat)
  
}

# Apply the QC function
hourly_count_pat <- QC_hourly_count(pat, minCount = 42)

# Examine the results
timeseriesTbl_multiPlot(hourly_count_pat$data, pattern = "pm25_[AB]")
```

In the code above, we calculate hourly aggregated average values, 
`mean_pat`, and hourly aggregated counts, `count_pat`. We use the counts 
to create a mask identifying hours where there is not enough data to meet
our minimum threshold, `minCount = 42`. The mean values for these hours are
replaced with `NA`.

Looking at the plot, we can verify that low count values are associatd with
missing values of `pm25_A/B`.

## Example 2 -- Z-score based QC

In this example, we construct a more complex QC algorithm using 
advanced features of `pat_aggregate()` and a coding style apporopriate for 
seasoned R developers.

The `FUN` parameter is defined "anonymously" (aka on-the-fly) and calculates a
Z-score for each 20-minute period in the data. If the Z-score is below 
our `0.05` threshold, we replace it with `NA`. The 20-minute data is then 
re-aggregated to an 1-hour average and returned. Effectively, this 
quality-control function is a basic two-tailed Z-score filter that only 
aggregates data if the it is within the p-value (0.05).

```{r}

QC_hourly_z_score <- function(pat, threshold = 0.05) {

  pat_z_scored_20min_avg <- 
    pat_aggregate(
      pat,
      function(x) {
        # Caluclate x z-scores
        z_score <- scale(na.omit(x)) # (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
        
        # Create z-score mask -
        # flag the z-scores who are out of the 5% and 95% quantile range
        z_mask <- pnorm(-abs(z_score)) < threshold
        x[z_mask] <- NA
        
        # Calculate the 20-min average values without the flagged data
        y <- mean(x, na.rm = TRUE)
        return(y)
      },
      unit = 'minutes',
      count = 20
    )
  
  # Re-aggregate to an 1-hour average of the data
  pat <- pat_aggregate(pat_z_scored_20min_avg)
  
  return(pat)
  
}

hourly_z_score_pat <- QC_hourly_z_score(pat)

# Examine the results
timeseriesTbl_multiPlot(hourly_z_score_pat$data, pattern = "pm25_[AB]")
```

-----

There are many ways to construct QC algorithms and the examples above illustrate 
what is possible within the *AirSensor* framework. We fullly expect people to
explore possibilities and create different QC algorithms depending on their
individual needs:

* How important is it to remove all potentially bad data?
* How important is it to retain all potentially good data?
* Are there specific QC steps that required by an organization?
* ...




