---
title: "MVCAA Tutorial 4: Exploring Airsensor Data"
author: "Mazama Science"
date: "2021-03-09"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MVCAA Tutorial 4: Exploring Airsensor Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 5)
```

## Introduction

This tutorial demonstrates how to analyze 'airsensor' data using functions
tailored to that data type, all of which have names beginning with `sensor_`. 
In order to run the code in this tutorial you must have 
followed the instructions in Tutorials 1-3 and created a directory with 'pas', 
'pat', and 'airsensor' data files for the Methow Valley. Target audiences 
include grad students, researchers and any member of the public
concerned about air quality and comfortable working with R and RStudio.

Tutorials in this series include:

* [MVCAA Tutorial 1: Creating PAT Data](https://mazamascience.github.io/AirSensor/articles/articles/MVCAA_Tutorial_1.html)
* [MVCAA Tutorial 2: Exploring PAT Data](https://mazamascience.github.io/AirSensor/articles/articles/MVCAA_Tutorial_2.html)
* [MVCAA Tutorial 3: Creating Airsensor Data](https://mazamascience.github.io/AirSensor/articles/articles/MVCAA_Tutorial_3.html)
* [MVCAA Tutorial 4: Exploring Airsensor Data](https://mazamascience.github.io/AirSensor/articles/articles/MVCAA_Tutorial_4.html)
* [MVCAA Tutorial 5: Building a Local Archive](https://mazamascience.github.io/AirSensor/articles/articles/MVCAA_Tutorial_5.html)
* [MVCAA Tutorial 6: Methow Vallely Smoke](https://mazamascience.github.io/AirSensor/articles/articles/MVCAA_Tutorial_6.html)

## Goal

Our goal in this tutorial is to explore a variety of `sensor_~()` functions available
in the **AirSensor** package and to explain their most important arguments and 
the interpretation of output graphics.

```{r setup-data-directory, eval = TRUE, warning = FALSE, message = FALSE}
# Methow Valley local data archive

# ----- Setup ------------------------------------------------------------------

library(AirSensor)

# Use the default archiveDir unless it is already defined
if ( !exists("archiveDir") ) {
  archiveDir <- file.path("~/Data/MVCAA")
}

# Load previously generated airsensor data
airsensorList <- get(load(file.path(archiveDir,"airsensorList.rda")))

# Print site names and associated ids
sapply(airsensorList, function(x) { return(x$meta$label) }) 

# Pull out each sensor as a separate 'airsensor' object
Balky_Hill <- airsensorList[["ab5dca99422f2c0d_13669"]]
Bush_School <- airsensorList[["49215ad49d1a87e3_10188"]]
Liberty_School <- airsensorList[["db5d6b3b79f5830e_39237"]]      # good
McFarland_Creek <- airsensorList[["f592adb5067ad9d3_13675"]]
Pine_Forest <- airsensorList[["0cbfeb2ce4c1553c_13661"]]         # great
Willowbrook_Farm <- airsensorList[["f96deab8c29aa42b_10134"]]    # poor
Benson_Creek <- airsensorList[["f6c44edd41c941c7_10182"]]
Gunn_Ranch <- airsensorList[["f736fd3fb21fc4da_13667"]]
Lower_Studhorse <- airsensorList[["4f19d256e1787973_10166"]]
Methow_Estates <- airsensorList[["4a47b9252e16e558_15077"]] 
Beaver_Creek <- airsensorList[["2e3b5ceea86a885b_10168"]]
Little_Cougar <- airsensorList[["96b108298883ca47_64441"]]
```

## sensor_calendarPlot

The [sensor_calendarPlot()](https://mazamascience.github.io/AirSensor/reference/pat_monitorComparison.html)
function plots PM2.5 concentration in a calendar format. This function is 
prepares data and ultimately calls the 
[openair](https://bookdown.org/david_carslaw/openair/) package
`calendarPlot()` function.

From the final plot in Tutorial 3, it is clear that the sensor labeled "Pine Forest" 
has the most complete data after hourly aggregation. Let's look at air quality 
impacts at "Pine Forest" using the 
[US EPA Air Quality Index colors](https://www.epa.gov/wildfire-smoke-course/wildfire-smoke-and-your-patients-health-air-quality-index)  
(`colors = "aqi"`). We expect that the PM2.5 values reported by the sensor will be exaggerated
to some extent. Nevertheless, the calendar plot will still show us which days during 
September 2020 had the worst PM2.5 levels, putting the health of Methow Valley 
residents at risk.

```{r sensor-calendarPlots, eval = TRUE, warning = FALSE, message = FALSE}
# Create a calendar plot for Pine_Forest
sensor_calendarPlot(
  sensor = Pine_Forest,
  colors = "aqi",
  title = "September 2020 -- Pine Forest"
)
```

From the calendar plot we see that the September 8-18 were all days of unhealthy
smoke levels with September 13-14 being the worst.

## sensor_polarPlot

The [sensor_polarPlot()](https://mazamascience.github.io/AirSensor/reference/sensor_polarPlot.html)
function wraps `openair::polarPlot()` which plots PM2.5 concentration as a 
function of wind speed and direction. This type of plot can be helpful in
determining the source of a pollutant.

If no wind data is specified, (default) `windData = NULL`, the function uses 
wind data from the nearest airport station obtained with the 
[worldmet](https://www.rdocumentation.org/packages/worldmet/versions/0.9.2/topics/worldmet) 
package. 

The color scale for some of the statistics is only to provide an indication of 
overall pattern and should not be interpreted in concentration units (µg/m3). 
(See the documentation for `openair::polarPlot()` for details.)

```{r sensor-polarPlot-Liberty_School, eval = TRUE, warning = FALSE, message = FALSE}
# Create a polar plot for Liberty School
polar_plot <- sensor_polarPlot(
  sensor = Liberty_School
)
```

The plot shows that there is evidence of increasing PM25 (smoke) concentrations 
at low wind speeds from all direction with more smoke coming from from the north.
You can use a function from the **PWFSLSmoke** package to create a local map where 
you will see that the Methow 
Valley runs primarily north-south at Liberty Bell High School.

```{r monitor_leaflet, eval = TRUE, warning = FALSE, message = FALSE}
PWFSLSmoke::monitor_leaflet(Liberty_School)
```

To see the wind speed/direction conditions that dominate the overall mean, we 
can use `statistic = "weighted_mean"`. As we can see from the plot below, wind 
from N with a speed between 6 and 4 m/s carries the highest PM25 concentration. 

This plot might suggest that wildfire smoke that entered the valley moved 
up and down the valley as part of typical, diurnal valley airflow patterns.

```{r sensor-polarPlot-Liberty_School-wm, eval = TRUE, warning = FALSE, message = FALSE}
# Create a polar plot for Pine Forest using the weighted mean
polar_plot <- sensor_polarPlot(
  sensor = Pine_Forest,
  statistic = "weighted_mean"
)
```

## sensor_pollutionRose

The [sensor_pollutionRose()](https://mazamascience.github.io/AirSensor/reference/sensor_pollutionRose.html)
function wraps `openair::pollutionRose()` which plots a traditional wind rose plot 
for wind direction and PM2.5. 

```{r sensor_pollutionRose_1, eval = TRUE, warning = FALSE, message = FALSE}
# Create a pollution rose plot for Liberty School using the proportion of the 
# frequency of measurements 
pollution_rose <- sensor_pollutionRose(
  sensor = Liberty_School
)
```

Highest PM25 concentrations are mostly associated with southerly winds. However 
the largest proportion of PM25 concentrations is associated with northerly winds.
By setting `statistic = "prop.mean"` , we can see that northerly winds 
contribute the most to overall concentrations as shown in the plot below. 

```{r sensor_pollutionRose_2, eval = TRUE, warning = FALSE, message = FALSE}
# Create a pollution rose plot for Liberty School using the relative contribution 
# of the measurements to the mean
sensor_pollutionRose(
  sensor = Liberty_School,
  statistic = "prop.mean"
)
```

More examples and explanations on how to interpret pollution rose plots can be 
found in 
["Section 5 Wind and Pollution Roses"](https://bookdown.org/david_carslaw/openair/sec-windRose.html) 
of The Openair Book. 

## sensor_extractData

The [sensor_extractData() and sensor_extractMeta() ](https://mazamascience.github.io/AirSensor/reference/sensor_extractDataFrame.html)
functions are convenient wrappers for extracting the airsensor `meta` and `data`
dataframes. These functions are designed to be useful when manipulating 
data in an R pipeline using `%>%`.

In the example below, we will use data manipulation functions from 
**PWFSLSmoke** and **dplyr** to calculate some daily statistics for Pine Forest.

```{r sensor_extractData, eval = TRUE, warning = FALSE, message = FALSE}
# Use MazamaCoreUtils::dateRange() to ensure we get POSIXct times in the local timezone
dateRange <- MazamaCoreUtils::dateRange(
  startdate = 20200906, 
  enddate = 20200920, 
  timezone = "America/Los_Angeles"
)
# Show daily max for Pine Forest  
Pine_Forest %>%
  PWFSLSmoke::monitor_subset(tlim = dateRange) %>%
  PWFSLSmoke::monitor_dailyStatistic(FUN = max) %>%
  sensor_extractData() %>%
  dplyr::mutate_if(is.numeric, round, 0)

# Show daily averages for Pine Forest 
Pine_Forest %>%
  PWFSLSmoke::monitor_subset(tlim = dateRange) %>%
  PWFSLSmoke::monitor_dailyStatistic(FUN = mean) %>%
  sensor_extractData() %>%
  dplyr::mutate_if(is.numeric, round, 0)

# Show daily standard deviations for Pine Forest  
Pine_Forest %>%
  PWFSLSmoke::monitor_subset(tlim = dateRange) %>%
  PWFSLSmoke::monitor_dailyStatistic(FUN = sd) %>%
  sensor_extractData() %>%
  dplyr::mutate_if(is.numeric, round, 1)
```


## PWFSLSmoke and AirMonitorPlots

All 'airsensor' objects are compatible with 'monitor' functions from the
[PWFSLSmoke](http://mazamascience.github.io/PWFSLSmoke/) 
package as well as functions from the
[AirMonitorPlots](http://mazamascience.github.io/AirMonitorPlots/) package for
generating **ggplot2** graphics.

The following examples are designed to encourage further exploration of those packages.

```{r monitor_ggTimeseries, eval = TRUE, warning = FALSE, message = FALSE}
library(PWFSLSmoke)
library(ggplot2)
library(AirMonitorPlots)

# "publication-ready" time-series plot
AirMonitorPlots::monitor_ggTimeseries(Pine_Forest)
```

```{r monitor_ggDailyHourlyBarplot, eval = TRUE, warning = FALSE, message = FALSE}
# Custom plot showing data + daily averages for the entire Valley
airsensorList %>%
  PWFSLSmoke::monitor_combine() %>%
  AirMonitorPlots::ggplot_pm25Timeseries() + 
  ggplot2::ggtitle("Methow Valley Smoke -- September, 2020") +
  AirMonitorPlots::geom_pm25Points(shape = "square", alpha = .1) + 
  AirMonitorPlots::stat_dailyAQCategory(alpha = .5) + 
  ggplot2::scale_y_continuous(limits = c(0, 300)) +
  AirMonitorPlots::custom_aqiStackedBar(width = 0.01) 
```

_Best of luck assessing air quality in your community!_
