---
title: "PurpleAir Synoptic Data"
author: "Mazama Science"
date: "2020-04-10"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PurpleAir Synoptic Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, message=FALSE}
library(AirSensor)
library(dplyr)
library(ggplot2)

knitr::opts_chunk$set(fig.width = 7, fig.height = 5)

# NOTE: Use example PAS data for vignettes to avoid long wait-times
data("example_pas")
pas <- example_pas
```

_Synoptic data_ provides a synopsis - a comprehensive view of something at a
moment in time. This vignette demonstrates an example 
workflow for exploring air quality synoptic data using the **AirSensor**
R package and data captured by [PurpleAir](https://www.purpleair.com/) air 
quality sensors.

## Synoptic Data Basics

### Creating Current Synoptic Data (slow)

PurpleAir sensor readings are uploaded to the cloud every 120 seconds.
(Every 80 seconds prior to a May 31, 2019 firmware upgrade.)
Data are processed by PurpleAir and a version of the data is displayed on the 
PurpleAir website. 

You can generate a current PurpleAir Synoptic (PAS) object (hereafter called
a `pas`) by using the `pas_createNew()` function. A `pas` object is just a large
dataframe with `r ncol(pas)` data columns and a record for each
PurupleAir sensor channel (2 channels per sensor).

The `pas_createNew()` function perfoms the following tasks under the hood: 

1. Download a raw dataset of the entire PurpleAir network that includes both 
metadata and recent PM2.5 averages for each deployed sensor across the globe. 
See `downloadParseSynopticData()` for more info.

2. Subset and enhance the raw dataset by replacing variables with more consistent,
human readable names amd adding spatial metadata for each sensor including the 
nearest official air quality monitor. For a more in depth explanation, see 
`enhanceSynopticData()`.

To create a new `pas` object you must first properly initialize the 
**MazamaSpatialUtils** package. The following example will create a brand new 
`pas` object with up-to-the-minute data:

_**NOTE: This can take up to a minute to process.**_

```{r pas_createNew, eval = FALSE}
library(AirSensor)

# Initialize spatial data processing 
library(MazamaSpatialUtils)
initializeMazamaSpatialUtils()

# Create a 'pas' object with current data
pas <- pas_createNew(countryCodes = "US")
```

### Loading Pre-generatted Synoptic Data (fast)

It is also possible to load pre-generated `pas` objects from a data archive.
These objects are updated regularly throughout each day and are typically used 
by other package functions primarily for the location metadata they contain. 
Archived `pas` objects from previous days will thus have data associated with 
near midnight of that date.

The archived `pas` objects can be loaded very quickly with the `pas_load()` 
function which obtains `pas` objects from the archive specified with
`setArchvieBaseUrl()`. When used without specifying the `datestamp` argument,
`pas_load()` will obtain the most recently processed `pas` object -- typically
less than an hour old.

```{r user_setup, eval = FALSE}
# Load packages
library(AirSensor)
library(dplyr)
library(ggplot2)

# Set location of pre-generated data files
setArchiveBaseUrl("https://airfire-data-exports.s3-us-west-2.amazonaws.com/PurpleAir/v1")

# Load the most recent archived 'pas' object
pas <- pas_load()
```

### PAS Data Structure

The `pas` dataset contains 45 columns, and each row corresponds to different 
PurpleAir sensors. For the data anlysis examples we will focus on the columns
labeled `stateCode`, `pm25_*`, `humidity`, `pressure`, `temperature`, and 
`pwfsl_closestDistance`. 

The complete list of columns is given below. Names in `ALL_CAPS` have been
retained from the PurpleAir .json file. Other columns have been renamed for
human readability.

```{r pas_names, echo = FALSE}
names(pas)
```

Let's take a quick peek at some of the PM2.5 data:

```{r kable}
# Extract and round just the PM2.5 data
pm25_data <-
  pas %>% 
  select(starts_with("pm25_")) %>% 
  round(1)

# Combine sensor label and pm2.5 data 
bind_cols(label = pas$label, pm25_data) %>%
  head(10) %>% 
  knitr::kable(
    col.names = c("label", "current", "10 min", "30 min", "1 hr", "6 hr", "1 day", "1 wk"),
    caption = "PAS PM2.5 Values"
  )
``` 

### Mapping `pas` PM2.5 Data

To visually explore a region, we can use our `pas` data with the `pas_leaflet()`
function to plot an interactive [leaflet](https://leafletjs.com/) map.
By default, `pas_leaflet()` will map the coordinates of each PurpleAir sensor 
and the hourly PM2.5 data. Clicking on a sensor will show sensor metadata.

```{r leaflet}
pas %>% 
  pas_leaflet()
```

If we want to narrow our selection, for example to California, we can look at which 
locations have a moderate to unhealthy 6-hour average air quality rating with
the following shor script that uses the `%>%` "pipe" operator:

```{r leaflet_pm25_6hr}
pas %>% 
  pas_filter(stateCode == 'CA') %>% 
  pas_filter(pm25_6hr >= 25.0) %>% 
  pas_leaflet(parameter = "pm25_6hr")
```

This code pipes our `pas` data into `pas_filter()` where we can set our 
selection criteria. The `stateCode` is the ISO 3166-2 state code, 
which tells `pas_filter()`  to subset for only those station sin California. 
The `pm25_6hr > 25.0` filter selects those records where the 6-hour average is 
above 25.0. The final funciton in the pipeline plots the remaining sensors
colored by `pm25_6hr`.

### Mapping `pas` Auxillary Data

We can also explore and utilize other PurpleAir sensor data. Check the 
`pas_leaflet()` documentation for all supported parameters.

Here is an example of humidity data captured from PurpleAir sensors across the 
state of California.

```{r leaflet_humidity}
pas %>% 
  pas_filter(stateCode == "CA") %>% 
  pas_leaflet(parameter = "humidity")
```

## Exploring PurpleAir Data

Because the `pas` object is a dataframe, we can use functionality from various 
other R packages. For example, `pas` data is compatible with "tidyverse" syntax 
so we can use **dplyr**, **ggplot2**, and **sf** to help transform, summarize, 
and visualize `pas` data in a _tidy_ way.

Below, we demonstrate the flexibility of **AirSensor**'s `pas` data objects and 
a few examples of exploratory data analysis pipelines, focusing on the state of 
Califronia and their PurpleAir sensor data. 

```{r init_explore, message=FALSE, warning=FALSE}
# Load the extra libraries
library(sf) 
library(ggplot2)

# Load spatial data with US counties
MazamaSpatialUtils::setSpatialDataDir('~/Data/Spatial/')
MazamaSpatialUtils::loadSpatialData("USCensusCounties")

# Subset to California and convert to an 'sf' spatial object
ca_counties <- 
  subset(USCensusCounties, stateCode == "CA") %>% 
  st_as_sf()

# Subset the 'pas' data to Californa and convert to `sf`
ca_pas <- 
  pas %>% 
  pas_filter(stateCode == 'CA') %>% 
  # Be sure to set the reference system 
  st_as_sf(
    coords = c('longitude', 'latitude'),  # identify coordinate variables
    crs = st_crs(ca_counties)             # set the reference system
  ) 
```

### County Mointor Counts

Using our California county spatial data, we can count the number of PurpleAir 
sensors that exist in the each region "feature" (aka polygon). This is an important metric to 
determine statistical confidence we might have -- _i.e._ the more sensors a region 
contains the more confidence we can have in any county-wide, aggregated statistic.  

(The `mutate()` function comes from **dplyr** while the `st_contains()` comes 
from **sf**. Some familiarity with these packages is required.)

```{r n_sensors, message=FALSE, warning=FALSE}
# Count the number of PurpleAir sensors in each County polygon feature
n_sensors <- 
  ca_counties %>% 
  mutate(sensorCount = lengths(st_contains(., ca_pas)))
```

Let's take a look at the counties that contain the most PurpleAir sensors.

```{r n_sensors_table, warning=FALSE, message=FALSE}
n_sensors %>% 
  select(sensorCount, name) %>% 
  filter(sensorCount > 50) %>%
  st_drop_geometry() %>% 
  arrange(desc(sensorCount)) %>% 
  knitr::kable(
    col.names = c("Sensor Count", "County"),
    caption = "Counties with >50 PurpleAir Sensors"
  )
```

#### Chloropleth Map

We can visualize the amount of PurpleAir sensors per county using the
**tmap** package. Here we create a "chloropleth" map, where each county is 
filled with a color representing the number of PurpleAir sensors it contains. 

```{r n_sensors_chloropleth}
library(tmap)

# Set up breaks
my_breaks = c(0,20,50,100,200,500,1000,Inf)
my_labels <- c("0 to 20", "20 to 50", "50 to 100", "100 to 200",
               "200 to 500", "500 to 1,000", "1,000 or more")

tm_shape(n_sensors) +
  tm_polygons(
    col = "sensorCount",
    palette = "YlOrBr",
    breaks = my_breaks
  )
```

#### Barplot

Sometimes a barplot is preferable for reading off values. Here is an 
example showing the top 15 counties with the the most PurpleAir sensors that are 
active.  

```{r n_sensors_barplot, fig.width = 5}
# Sort the counties by the number of PurpleAir sensors
sorted_n_sensors <- 
  n_sensors %>% 
  arrange(desc(sensorCount))

ggplot(sorted_n_sensors) + 
  geom_col(aes(
    x = reorder(name, sensorCount), 
    y = sensorCount, 
    fill = cut_number(sensorCount, n = 8, breaks = my_breaks, labels = my_labels)
  )) + 
  scale_fill_brewer(
    palette = "YlOrBr", 
    direction = 1
  ) +
  labs(fill = "Sensor Count") +
  xlab('County') + 
  theme_minimal() + 
  coord_flip() 

```

#### Region Averaged Sensor Data

Our `pas` object contains the latest ~2-minute resolution PM2.5 data 
(`pm25_current`) as well as other, longer interval averages of PM2.5, 
_e.g._ `pm25_1week`, which we can use to calculate region averaged weekly PM2.5 
sensor data.  

```{r mean_pm25, warning=FALSE, message=FALSE}
# Join the PurpleAir and county data and calculate the mean per county
mean_pm25 <- 
  ca_counties %>%
  st_join(ca_pas) %>% 
  group_by(name) %>% 
  summarise(pm25_weeklyMean = round(mean(pm25_1week, na.rm = TRUE), 1))

mean_pm25 %>% 
  select(name, pm25_weeklyMean) %>% 
  st_drop_geometry() %>% 
  arrange(desc(pm25_weeklyMean)) %>% 
  head(20) %>%
  knitr::kable(
    col.names = c("County", "Mean Sensor PM2.5"),
    caption = "Counties with Highest weekly PM2.5"
  )
```

It is important to clarify that this data **_does not_** represent a region's 
"spatially averaged" air quality. It is only a sipmle summary of data from the 
avaliable PurpleAir sensors without any regard for how they are spatially
distributed.

### Chloropleth 

In a similar fashion to that illustrated above, we can also create a 1-week 
PurpleAir averages per county chloropleth map.  

```{r mean_pm25_chloropleth, warning=FALSE, message=FALSE}
# Set up breaks
my_breaks = seq(0,40,5)
my_labels <- paste(my_breaks[1:8], my_breaks[2:9], sep = " to ")

tm_shape(mean_pm25) +
  tm_polygons(
    col = "pm25_weeklyMean",
    palette = "YlOrBr",
    breaks = seq(0,40,5)
  )
```

#### Barplot

Just as before, a barplot is sometimes a better choice. Below is an example of 
plotting the Top 15 1-week PurpleAir county averages.

```{r mean_pm25_barplot, fig.width = 5, warning=FALSE, message=FALSE}
# Sort the weekly averages per county and select the top 15
top15_mean_pm25 <- 
  mean_pm25 %>% 
  arrange(desc(pm25_weeklyMean)) %>% 
  head(n = 15)

# NOTE:  The color legend is not identical to the legend above becuase we are
# NOTE:  only plotting a subset of all counties.

ggplot(top15_mean_pm25) + 
  geom_col(aes(
    x = reorder(name, pm25_weeklyMean), 
    y = pm25_weeklyMean, 
    fill = cut_number(pm25_weeklyMean, n = 8, breaks = my_breaks, labels = my_labels)
  )) +
  scale_fill_brewer(
    palette = "YlOrBr", 
    direction = 1
  ) +
  labs(fill = "Sensor Count") +
  xlab('County') + 
  theme_minimal() + 
  coord_flip() 
```

_Happy Exploring!_

----

_Mazama Science_


