---
title: "PurpleAir Synoptic Data"
author: "Mazama Science"
date: "2020-04-10"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PurpleAir Synoptic Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, message=FALSE}
library(AirSensor)
library(dplyr)
library(ggplot2)

knitr::opts_chunk$set(fig.width = 7, fig.height = 5)

# NOTE: Use example PAS data for vignettes to avoid long wait-times
data("example_pas")
pas <- example_pas
```

_Synoptic data_ provides a synopsis - a comprehensive view of something at a
moment in time. This vignette demonstrates an example 
workflow for exploring air quality synoptic data using the **AirSensor**
R package and data captured by [PurpleAir](https://www.purpleair.com/) air 
quality sensors.

## Synoptic Data Basics

### Creating Current Synoptic Data (slow)

PurpleAir sensor readings are uploaded to the cloud every 120 seconds.
(Every 80 seconds prior to a May 31, 2019 firmware upgrade.)
Data are processed by PurpleAir and a version of the data is displayed on the 
PurpleAir website. 

You can generate a current PurpleAir Synoptic (PAS) object (hereafter called
a `pas`) by using the `pas_createNew()` function. A `pas` object is just a large
dataframe with `r ncol(pas)` data columns and a record for each
PurupleAir sensor channel (2 channels per sensor).

The `pas_createNew()` function performs the following tasks under the hood: 

1. Download a raw dataset of the entire PurpleAir network that includes both 
metadata and recent PM2.5 averages for each deployed sensor across the globe. 
See `downloadParseSynopticData()` for more info.

2. Subset and enhance the raw dataset by replacing variables with more consistent,
human readable names and adding spatial metadata for each sensor including the 
nearest official air quality monitor. For a more in depth explanation, see 
`enhanceSynopticData()`.

To create a new `pas` object you must first properly initialize the 
**MazamaSpatialUtils** package. The following example will create a brand new 
`pas` object with up-to-the-minute data:

_**NOTE: This can take up to a minute to process.**_

```{r pas_createNew, eval = FALSE}
library(AirSensor)
library(dplyr)
library(ggplot2)

# Define PURPLE_AIR_API_READ_KEY in a .gitignore protected file
source("global_vars.R")
setAPIKey("PurpleAir-read", PURPLE_AIR_API_READ_KEY)

# Initialize spatial data processing 
library(MazamaSpatialUtils)
initializeMazamaSpatialUtils()

# Create a 'pas' object with current data
pas <- 
  pas_createNew(
    countryCodes = "US", 
    stateCodes = "CA", 
    counties = "Los Angeles"
  )
```

### Loading Pre-generated Synoptic Data (fast)

It is also possible to load pre-generated `pas` objects from a data archive.
These objects are updated regularly throughout each day and are typically used 
by other package functions primarily for the location metadata they contain. 
Archived `pas` objects from previous days will thus have data associated with 
near midnight of that date.

The archived `pas` objects can be loaded very quickly with the `pas_load()` 
function which obtains `pas` objects from the archive specified with
`setArchvieBaseUrl()`. When used without specifying the `datestamp` argument,
`pas_load()` will obtain the most recently processed `pas` object -- typically
less than an hour old.

```{r user_setup, eval = FALSE}
# Load packages
library(AirSensor)
library(dplyr)
library(ggplot2)

# Set location of pre-generated data files
setArchiveBaseUrl("https://airfire-data-exports.s3-us-west-2.amazonaws.com/PurpleAir/v1")

# Load the most recent archived 'pas' object
pas <- pas_load()
```

### PAS Data Structure

The `pas` dataset contains 45 columns, and each row corresponds to different 
PurpleAir sensors. For the data analysis examples we will focus on the columns
labeled `stateCode`, `pm25_*`, `humidity`, `pressure`, `temperature`, and 
`pwfsl_closestDistance`. 

The complete list of columns is given below. Names in `ALL_CAPS` have been
retained from the PurpleAir .json file. Other columns have been renamed for
human readability.

```{r pas_names, echo = FALSE}
names(pas)
```

Let's take a quick peek at some of the PM2.5 data:

```{r kable}
# Extract and round just the PM2.5 data
pm25_data <-
  pas %>% 
  select(starts_with("pm2.5_")) %>% 
  round(1)

# Combine sensor label and pm2.5 data 
bind_cols(label = pas$locationName, pm25_data) %>%
  head(10) %>% 
  knitr::kable(
    col.names = c("location name", "10 min", "30 min", "1 hr", "6 hr", "1 day", "1 wk"),
    caption = "PAS PM2.5 Values"
  )
``` 

### Mapping `pas` PM2.5 Data

To visually explore a region, we can use our `pas` data with the `pas_leaflet()`
function to plot an interactive [leaflet](https://leafletjs.com/) map.
By default, `pas_leaflet()` will map the coordinates of each PurpleAir sensor 
and the hourly PM2.5 data. Clicking on a sensor will show sensor metadata.

```{r leaflet}
pas %>% 
  pas_leaflet(parameter = "pm2.5_60minute")
```

If we want to narrow our selection, for example to California, we can look at which 
locations have a moderate to unhealthy 6-hour average air quality rating with
the following short script that uses the `%>%` "pipe" operator:

```{r leaflet_pm2.5_6hour}
pas %>% 
  pas_filter(pm2.5_6hour >= 15.0) %>% 
  pas_leaflet(parameter = "pm2.5_6hour")
```

This code pipes our `pas` data into `pas_filter()` where we can set our 
selection criteria. The `stateCode` is the ISO 3166-2 state code, 
which tells `pas_filter()`  to subset for only those station sin California. 
The `pm25_6hr > 25.0` filter selects those records where the 6-hour average is 
above 25.0. The final function in the pipeline plots the remaining sensors
colored by `pm25_6hr`.

### Mapping `pas` Auxiliary Data

We can also explore and utilize other PurpleAir sensor data. Check the 
`pas_leaflet()` documentation for all supported parameters.

Here is an example of humidity data captured from PurpleAir sensors across the 
state of California.

```{r leaflet_humidity}
pas %>% 
  pas_leaflet(parameter = "humidity")
```


_Happy Exploring!_

----

_Mazama Science_


