---
title: "An Introduction to PurpleAir TimeSeries Data"
author: "Mazama Science"
date: "2019-04-19"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An Introduction to PurpleAir TimeSeries Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(fig.width=7, fig.height=5)
library(MazamaCoreUtils)
library(AirSensor)
library(dplyr)
library(ggplot2)
# NOTE: Use example PAS and PAT data for vignettes to avoid long wait-times
pas <- AirSensor::example_pas
pat <- AirSensor::example_pat
```

_Time series_ data provides a minute-by-minute database structure for transforming 
and analyzing [PurpleAir](https://www.purpleair.com/) sensor data.
This vignette demonstrates an example analysis of an individual monitor located
in Seattle, Washington over a two-month duration in which the Pacific Northwest 
experienced [hazardous air-quality conditions](https://www.usnews.com/news/healthiest-communities/articles/2018-08-19/air-quality-to-worsen-in-northwest-as-smoke-returns) caused by wildfires in British Columbia, Canada. 

**Disclaimer:** It is highly recommended that you read `vignettes/pas_introduction.Rmd`
before beginning this tutorial.

## Loading PurpleAir Timeseries data
PurpleAir sensor readings are uploaded to the cloud every 120 seconds 
where they are stored for download and display on the PurpleAir website. After
every interval, the synoptic data is refreshed and the outdated synoptic data is 
then stored in a [ThingSpeak](https://thingspeak.com) database. In order to 
access the ThingSpeak channel API we must first load the synoptic database, 
which provides access tokens for each PurpleAir sensor -- accessed by the 
monitor of interests label. 

```
library(MazamaCoreUtils)
library(AirSensor)
pas <- pas_load()
pat <- pat_createNew(pas, "Seattle", startdate = 20180701, enddate = 20180901)
```

Notice that when passing our synoptic dataframe "`pas`" to `pat_createNew()`, 
we also supply a unique identification label and date-interval. In this case, our 
monitor-of-interest (MOI) is a sensor named "Seattle" and our dates-of-interest
are 2018-07-01 to 2018-09-01. 

**Note:** You must provide a label to `pat_createNew()` in 
order to supply ThingSpeak with the necessary metadata to access the sensors 
database. A time range on the other hand is optional; by default not providing 
`pat_createNew()` with a start and end date will return  the most recent week 
of time series data.

## The PurpleAirTimeseries ("pat") Data Model
Let's begin by exploring the attributes of the dataframe returned by the 
`pat_createNew()` function. 

```{r names}
pat %>%
  names()
```

`pat` contains _two_ dataframes, `meta` and `data`. 

The `meta` dataframe contains metadata of the selected PurpleAir sensor -- 
this includes non-time series data such as location information, labels, etc. 
The `data` dataframe contains datestamped sensor readings of 
PM2.5, temperature, humidity, and other pertinent sensor data. 

We'll start by plotting PurpleAir's raw sensor data. We can quickly display the 
time series data by using `pat_multiplot()` and passing in our raw `pat` and
desired plot type ("`all`" sensor data). 

```{r allPlot}
pat %>%
  pat_multiplot(plottype = "all")
```

## Exploring Time Series Data

Our `pat` dataframe spans two months. While this provides a great overview of
PM2.5, it is unwieldy to analyze if we are only interested in anomalous air 
quality. We can use `pat_filterDate()` to subset our `pat` dates. In this case,
we'll reduce our time range to 2018-08-01 - 2018-09-01.

```{r augustPlot}
pat_august <- 
  pat %>% 
  pat_filterDate(startdate = 20180801, enddate = 20180901)

pat_august %>%
  pat_multiplot(plottype = "pm25_over")
```

We can look for correlations in the raw data with `pat_scatterplot()`. When a
sensor is properly functioning, the only correlations will be a strong positive
one between between the A and B channels (`pm25_A:pm25_B`) and a strong
negative one between temperature and humidity.

```{r scatterplot, message = FALSE, warning = FALSE}
  pat_august %>%
    pat_scatterplot()
```

## Outlier Detection

Our `pat_august` "pat" object displays some intermittent sensor errors that
appear as spikes in the data. In order to identify and remove  PM2.5 outliers 
like these we can use `pat_outliers()`. By default, this function will create a
plot of the raw data with outliers marked with a red asterisk. It can also be
used to replace outliers with window median values.

```{r outlierPlot}
pat_august_filtered <- 
  pat_august %>%
  pat_outliers(replace = TRUE, showPlot = TRUE)
```

## A/B Channel Comparison

Now that we have a filtered dataset we can subset the data further and 
examine it in more detail. The `pat_internalFit()` function will compare PM2.5 
data from the A and B channels to verify that the sensor is functioning
properly.

```{r internalFit, message = FALSE, warning = FALSE}
one_week <- 
  pat_august %>% 
  pat_filterDate(startdate = 20180813, days = 7)

# Channel A/B comparison
one_week %>% 
  pat_internalFit()
```

The high R^2^ value indicates that the two channels are highly correlated while a slope
of ~0.9 suggests a slight relative bias in the measurements. (Perfect alignment
would have a slope of 1.0.)

## Comparison with Federal Monitors

For locations near federal monitors that are part of the 
[USFS Monitoring site](https://tools.airfire.org/monitoring/v4), we can also
compare the sensor data with hourly data from a federal monitor

```{r externalFit, message = FALSE, warning = FALSE}
# Sensor/Monitor comparison
one_week %>% 
  pat_externalFit()
```

Overall, this is an excellent fit with the PurpleAir sensor capturing the 
temporal evolution of the wildfire smoke event impacting Seattle. The sensor
data is biased a little high relative to the monitoring data but the much 
higher temporal resolution of the sensor provides a rich dataset to work with.

This package contains many additional functions for working with PurpleAir 
data and users are encouraged to review the reference documentation.

_Happy Exploring!_

_Mazama Science_

