---
title: "PurpleAir Failure Modes"
author: "Mazama Science"
date: "7/09/2019"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PurpleAir Failure Modes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette examines different failure modes we have found in the data from
PurpleAir (PA-II) sensors.

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=7, fig.height=5)
library(AirSensor)
setArchiveBaseUrl("http://smoke.mazamascience.com/data/PurpleAir")
pas <- pas_load(archival = TRUE)
```

## Channel A Moderate Noise

The A channel readings are moderately more noisy than those of the B channel.

```{r channel_a_noise_1, echo=FALSE, message=FALSE, warning=FALSE}
pat_a_noisy_1 <- example_pat_failure_A
pat_internalFit(pat_a_noisy_1)
```

Channel A here often measures above B at the same moment in time, as shown by
the scatterplot with the vertical clump near the origin instead of along the 
angled regression line.

## Channel A Extreme Noise

Channel A readings show extreme levels of noise.

```{r channel_a_noise_2, echo=FALSE}
pat_a_noisy_2 <- pat_createNew(pas, 
                               label = "SCAP_14", 
                               startdate = 20190701, 
                               enddate = 20190708,
                               timezone = "America/Los_Angeles")
pat_multiplot(pat_a_noisy_2, plottype = "pm25_over")
```

While a small amount of noise is natural when measuring particulate matter, 
sometimes the noise level goes far beyond what is allowable. Here we see the A 
channel looks like a cloud of points compared to the much more consistent 
channel B. A faint wave pattern can still be identified, however.

## Channel A Jumps

Channel A shows a sudden, but short-lived jump in PM2.5 readings.

```{r channel_a_jump, echo=FALSE}
pat_a_jump <- example_pat_failure_B
pat_multiplot(pat_a_jump, plottype = "pm25_over")
```

This plot shows a jump that seems to retain a consistent wave pattern
for a while rather than just being random noise. This could possibly be a 
temporary mix of the "Matches Humidity" failure mode.

## Channel A Matches Humidity

The A channel PM2.5 sensor starts reflecting humidity readings instead of PM 
levels.

```{r channel_a_humidity_multi, echo=FALSE, warning=FALSE}
pat_a_humidity <- pat_createNew(pas, "BikeSGV - West Pasadena", 
                                startdate = "2019-04-16", 
                                enddate = "2019-04-24",
                                timezone = "America/Los_Angeles")
pat_multiplot(pat_a_humidity)
```

You can see in the multiplot the clear disconnect between the two PM2.5 
channels. Both sensors appear to agree with each other until channel A suddenly 
jumps into the thousands and starts tracing the trend of the humidity data 
(plotted directly below).

```{r, channel_a_humidity_scatter, echo=FALSE, warning=FALSE}
pat_scatterplot(pat_a_humidity)
```

A glance at these scatterplots gives a us another look at just how uncorrelated 
the A and B channels are, while the relationships for channel A 
with temperature and humidity are abnormally well-defined. It's actually not
very clear which of the auxillary sensors the A channel is reflecting since
temperature and relative humidity are naturally correlated themselves.

A more in-depth analysis of this issue is provided in 
`local_examples/bikesgv_story.Rmd` .

## Channel A Magic Number

The A channel is centered around a particular level but is sometimes affected
by humidity when it goes past a certain threshold.

```{r, echo=FALSE, warning=FALSE}
pat <- pat_createNew(pas, "SCEM_05", 
                     startdate = 20190701, 
                     enddate = 20190710,
                     timezone = "America/Los_Angeles")
pat_multiplot(pat)
```

The multiplot shows the A and B channels have very different readings, similar 
to the "Matches Humidity" failure mode. The strange thing here though is that 
the A channel is mostly flat, with only the occasional spike when the humidity 
measures very high. Let's see what number it is centered around:

```{r}
plot(pat$data$datetime, pat$data$pm25_A, 
     ylim = c(3325, 3340), 
     pch = 15, cex = 0.6, col=adjustcolor("black", 0.2),
     xlab = "2019", ylab = "PM2.5 A")

temp <- table(as.vector(pat$data$pm25_A))
print(paste0("Mode value: ", names(temp)[temp == max(temp)]))
```

Although there is plenty of noise between 3325 and 3340 ug/m3, a very clear, 
very straight line of points is visible at exactly 3333.0 ug/m3.

## Channel B is Zero

Channel B measures no particulate matter at all.

```{r channel_b_zero, echo=FALSE}
pat_b_zero <- pat_createNew(pas, "SCAP_46", 
                            startdate = "2019-07-01", 
                            enddate = "2019-07-08",
                            timezone = "America/Los_Angeles")
pat_multiplot(pat_b_zero, sampleSize = NULL)

simple <- dplyr::select(pat_b_zero$data, datetime, pm25_A, pm25_B)
head(simple)
```

In this case one may wish to work with the A channel data only
