---
title: "Temporally Aggregated Time Series"
author: "Mazama Science"
date: "2020-05-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Temporal Aggregation

An essential piece of analysis for large data sets is efficient granularization: 
computing aggregations like `sum`, `mean`, `sd`, `min`, and `max`, in which a 
single number gives insight into the nature of a larger population of measurements. 
_Time series_ aggregation is the aggregation of all data points over a specified 
_period_.  Within the **AirSensor** package, this is achieved with `pat_aggregate()` 
which applies an aggregating function, similar to those mentioned above, over a 
temporal subset of data. By default time series data is broken up into 1-hour 
periods.  The result of the aggregation is a new dataset where each data point reflects a 
statistical view of the collected and aggregated data points over each hour.

To demonstrate this feature we'll load a 24-hour period of Purple Air data and 
compare the data to the aggregated data.

```{r load_data}
# AirSensor setup
library(AirSensor)
setArchiveBaseUrl("http://data.mazamascience.com/PurpleAir/v1")

# Load the PurpleAir sensor data
pas <- pas_load(archival = TRUE)
pat <- 
  pat_load(
    label = 'SCSC_33', 
    pas = pas,
    startdate = 20200501,
    enddate = 20200502
  )
```

### No Aggregation

A standard 24-hour period of non-aggregated data typically consists of 720 data 
entries -- one record every 2 minutes.

```{r}
nrow(pat$data)
```

In the multi-plot below we can see the high temporal resolution of the raw data.

```{r raw_multiplot}
pat_multiPlot(pat, sampleSize = NULL)
```

### Hourly Aggregation

Using `pat_aggregate()` we can aggregate the `pat` object to an hourly average 
of the data. Hourly reporting is the standard for most regulatory air quality
monitoring and is the recommended _period_ to use. It is also the default.

```{r aggregate}
hourly_pat <- pat_aggregate(pat)
nrow(hourly_pat$data)
```

As we'd expect, an hourly aggregated `pat` contains 24 records, one for each hour.
Spikes seen in the raw data contribute to each hourly average but the overall
effect is a much smoother time series.

```{r hourly_multiplot}
pat_multiPlot(hourly_pat)
```

### Sub-hour Aggregation

With care, we can extend the use of `pat_aggregate()` to summarize time series 
`pat` data for nearly _any_ period. Sub-hour aggregation may be useful in 
creating custom QC functions.

You can create different aggregation periods by explicitly providing `unit`, a 
string describing the period to split by `unit =` `'hours'`, `'minutes'`, 
`'weeks'`, `'months'`, _etc._ and `count`, the number of untis to aggregate in each 
bin. For example, a 15-minute standard deviation (`sd`) aggregation would look 
like this:

```{r aggregate_15}
# Aggregate the standard deviation of 15-minute periods
sd_fifteen_minute_pat <- 
  pat_aggregate(pat, 
                function(x) { sd(x, na.rm = TRUE) }, 
                unit = 'minutes', 
                count = 15)

# View first 5 entries of data
head(sd_fifteen_minute_pat$data)
```

### Custom Aggregation Functions

In order to write custom aggregation functions for use with `pat_aggregate()`, 
we must first familiarize ourselves with `pat_aggregate()`'s underlying 
algorithm.

When executed, `pat_aggregate(pat, FUN)` utilizes the `datetime` axis 
of a PurpleAir Timeseries object (`pat`) to split the data into time-granular 
bins. For each column of numeric data within `pat$data`, `pat_aggregate()`
applies `FUN` to the binned data to produce an hourly (by default) vector
of values.

`FUN` may be assigned to _any_ valid R function! _(Caveat: With great power) 
comes great responsibility.)_ The only requirements for 
the function are that `FUN` _must_ operate on univariate numeric data and return a 
scalar value (think `sum`, or `mean`). The last step in `pat_aggregate()` is to 
combine the transformed bins along a similarly binned `datetime` axis and return 
a data object of the same `pat` class (`pa_timeseries`).

The ability to create custom functions for use in aggregation opens the door
wide for exploratory data analysis and QC design.

