---
title: "Working With Wind Data"
author: "Mazama Science"
date: "2019-07-24"
output: html_document
vignette: >
  %\VignetteIndexEntry{Working With Wind Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(AirSensor)
library(dplyr)
library(worldmet)
```

Wind speed and direction are required to generate rose plots or polar plots for 
PM2.5. Loading this data is made easy using the `worldmet` R package which makes
accessible over 35,000 NOAA measurement sites, many of which are located close 
to South Coast monitors.

In this example we'll be generating plots for Seal Beach during the period 
between July 1st and 7th, 2019. We'll first need to locate a NOAA site that is 
reasonably close to our sensor, and this can be quickly done by looking at a map
of sites from the `worldmet::getMeta()` function:

```{r show_site_map}
worldmet::getMeta(lon = -118, lat = 34, n = 25, returnMap = TRUE)
```

Now we can zoom in to Long Beach to find our communtity and a nearby NOAA site. 
The "LOS ALAMITOS AAF AIRPORT" looks like the closest choice to Leisure 
World--just about 4km away. Clicking on its icon shows site information 
including its unique code `722975-53141`.

Now we can use the `worldmet::importNOAA()` function to load in all of the site
readings by providing the site code we just found and the year of interest.

(By default this function runs in parallel, but this parameter must be set to 
false to run in the vignette)

```{r load_site_data, warning=FALSE}
# Always specify a timezone wherever possible!
timezone <- "America/Los_Angeles"

# Define the timeframe.
start <- lubridate::ymd("2019-06-01", tz = timezone)
end   <- lubridate::ymd("2019-06-30", tz = timezone)

# Load Los Alamitos Airport site data for the timeframe.
siteData <- worldmet::importNOAA(code = "722975-53141",
                                 year = 2019,
                                 parallel = FALSE)

names(siteData)
```

The rose plot and polar plot functions require a dataframe with only `date`, 
`wd`, and `ws`, so let's subset these site readings for just the variables we 
need in our timeframe (not the whole year!):

```{r subset_wind_data, echo=TRUE}
windData <- 
  dplyr::select(siteData, c("date", "wd", "ws")) %>% 
  dplyr::filter(date >= start, date < end)
```

And now we can generate rose and polar plots with our wind data and a sensor
object in the same timeframe.

```{r plot_data, echo=TRUE, warning=FALSE}
setArchiveBaseUrl("http://smoke.mazamascience.com/data/PurpleAir")

sensor <- 
  AirSensor::sensor_load(startdate = start, enddate = end) %>%
  AirSensor::sensor_filterMeta(monitorID == "SCSB_02")

AirSensor::sensor_pollutionRose(sensor, windData, statistic = "prop.mean")
AirSensor::sensor_polarPlot(sensor, windData, resolution = "normal")
```